Genetic evidence: zebrafish hoxba and hoxbb clusters are essential for the anterior-posterior positioning of pectoral fins

Version 1

Abstract

Abstract

Vertebrate paired appendages, such as the pectoral fins in fish and the forelimbs in tetrapods, emerge at specific regions along the anterior-posterior axis of the body. Hox genes are considered prime candidates for determining the positioning of these paired appendages during development. Despite extensive phenotypic analyses of numerous single and compound Hox knockout mice, no genetic studies have identified substantial defects in limb positioning, leaving questions unresolved. In a previous study, we generated seven distinct hox cluster-deficient mutants in zebrafish. Here, we provide genetic evidence that zebrafish hoxba;hoxbb cluster-deleted mutants specifically exhibit a complete absence of pectoral fins, accompanied by the absence of tbx5a expression in pectoral fin buds. In these mutants, tbx5a expression in the pectoral fin field of the lateral plate mesoderm fails to be induced at an early stage, suggesting a lack of pectoral fin precursor cells. Furthermore, the competence to respond to retinoic acid is lost in hoxba;hoxbb cluster mutants, indicating that tbx5a expression cannot be induced in the pectoral fin buds. We also identify hoxb4a, hoxb5a, and hoxb5b as pivotal genes underlying this process. Although the frameshift mutations in these hox genes do not recapitulate the absence of pectoral fins, we demonstrate that deletion mutants at these genomic loci show the absence of pectoral fins, albeit with low penetrance. Our results suggest that the positioning of zebrafish pectoral fins is cooperatively determined by hoxb4a, hoxb5a, and hoxb5b within hoxba and hoxbb clusters, which induce tbx5a expression in the restricted pectoral fin field. Our findings also provide insights into the acquisition of paired appendages in vertebrates.

Introduction

In jawed vertebrates, paired appendages—such as the pectoral and pelvic fins in fish, and their homologous forelimbs and hindlimbs in tetrapods—develop at precise locations along the anterior-posterior axis of each species. These paired appendages arise from progenitor cells located in distinct regions of the lateral plate mesoderm (Murata et al., 2010; Nishimoto and Logan, 2016; Shimada et al., 2013). The positioning of these paired appendages has long fascinated researchers, yet our understanding of the molecular mechanisms underlying this process remains limited.

In bilaterian animals, Hox genes—encoding evolutionarily conserved homeodomain-containing transcription factors—provide positional information and developmental timing along the anterior-posterior axis (Iimura and Pourquie, 2007; Izpisua-Belmonte and Duboule, 1992; Krumlauf, 1994). A defining feature of Hox genes is their structural organization into Hox clusters, where multiple Hox genes are arranged in a precise order. Additionally, Hox clusters exhibit a distinctive phenomenon known as Hox collinearity, where the genomic arrangement of Hox genes correlates with specific developmental regions along the body axes (Dolle et al., 1989; Duboule and Dolle, 1989; Graham et al., 1989). In vertebrates, Hox clusters underwent divergence due to two rounds of whole-genome duplication early in vertebrate evolution (Dehal and Boore, 2005; Ohno, 1970), leading to the establishment of four distinct Hox clusters (HoxA, HoxB, HoxC, and HoxD), each consisting essentially of 1-13 paralogous groups in tetrapods. In contrast, teleost fishes experienced an additional teleost-specific whole-genome duplication, followed by the loss of a hox cluster, resulting in seven hox clusters in zebrafish (Amores et al., 1998; Woltering and Durston, 2006).

Previous studies in mice and chickens suggest that the initial position of limbs is likely regulated by Hox genes (Tanaka, 2013; Tickle, 2015). Supporting this hypothesis, earlier research has shown that the anterior boundaries of specific Hox gene expression domains align with the future positions of limbs (Burke et al., 1995; Cohn et al., 1997). Moreover, experiments with avian embryos, in which several Hox genes were overexpressed or functionally interfered with, demonstrated altered positions of forelimb buds (Moreau et al., 2019). Additionally, it has been established that Hox proteins bind to the enhancer of Tbx5 gene, which is crucial for the initial formation of forelimbs, thereby inducing Tbx5 expression in the limb buds (Minguillon et al., 2012; Nishimoto et al., 2014). However, despite the generation of numerous knockout mice for single and multiple Hox genes, Hoxb5 knockout mice exhibit only a rostral shift in the position of forelimb buds, with incomplete penetrance (Rancourt et al., 1995), and no significant abnormalities have been observed in the early stages of limb formation. This lack of genetic evidence for limb positioning by Hox genes sharply contrasts with results from genetic analyses of various knockout mice, which have shown that the paralogous 9-13 Hox genes in HoxA and HoxD clusters are cooperatively essential for proximal-distal patterning of limbs after limb bud formation (Boulet and Capecchi, 2004; Davis et al., 1995; Fromental-Ramain et al., 1996a; Fromental-Ramain et al., 1996b; Kmita et al., 2005). Consequently, the precise role of Hox genes in defining the initial position of limb formation along the anterior-posterior axis remains unclear.

Using zebrafish, which belong to the Actinopterygii class, distinct from the tetrapods of the Sarcopterygii class, we previously generated mutants lacking each of the seven zebrafish hox clusters using the CRISPR-Cas9 method (Yamada et al., 2021). In our recent genetic analyses of hoxaa, hoxab, and hoxda clusters—homologous to the mouse HoxA and HoxD clusters—we demonstrated that, similar to mice, these zebrafish hox clusters cooperatively play an essential role in the formation of the pectoral fins (Ishizaka et al., 2024), which are homologous to the forelimbs. In this study, we provide the first genetic evidence, to our knowledge, that Hox genes specify the positions of paired appendages in the vertebrates. The double-deletion mutants of hoxba and hoxbb clusters, derived from the ancient HoxB cluster, exhibit a complete absence of pectoral fins due to the failure to express tbx5a. Despite incomplete penetrance, we propose a model in which hoxb4a, hoxb5a, and hoxb5b cooperatively provide positional cues along the anterior-posterior axis within the lateral plate mesoderm, thereby specifying the initial positions for fin bud formation through the induction of tbx5a in pectoral fin field.

Results

The absence of pectoral fins in zebrafish hoxba;hoxbb cluster-deleted embryos

In a previous study, we created seven individual hox cluster-deficient mutants in zebrafish using the CRISPR-Cas9 system (Yamada et al., 2021). Among these mutants, hoxba cluster-deleted embryos exhibited morphological abnormalities in their pectoral fins at 3 dpf (Figure 1A-C). To further investigate the pectoral fin phenotype in hoxba cluster mutants, we first analyzed the expression patterns of tbx5a, which plays a key role in the initial induction of pectoral fin buds in zebrafish (Ahn et al., 2002; Garrity et al., 2002; Ng et al., 2002). When comparing with tbx5a expression in wild-type embryos, we found that the tbx5a signal was reduced in the pectoral fin buds of hoxba cluster mutants (Figure 1H, I). Given that hoxba and hoxbb clusters originated from the ancestral HoxB cluster through teleost-specific whole-genome duplication (Amores et al., 1998), there may be functional redundancy between them. Surprisingly, we discovered that the simultaneous deletion of both hoxba and hoxbb clusters resulted in the complete absence of pectoral fins (Figure 1A, G). In contrast, pectoral fins were present in hoxba-/-;hoxbb+/- and hoxba+/-;hoxbb-/- mutants (Figure 1E, F), indicating that an allele from either hoxba or hoxbb cluster is sufficient for the pectoral fin formation. Moreover, all embryos lacking pectoral fins were identified as hoxba;hoxbb double homozygous mutants, with the expected penetrance (n=15/252; 5.9%) being consistent with predictions based on Mendelian genetics (1/16=6.3 %). Furthermore, while hoxba;hoxbb cluster double homozygous mutants are embryonic lethal around 5 dpf, we could not detect any trace of pectoral fin development. Alongside this phenotype, the expression of tbx5a was significantly reduced to nearly undetectable levels in hoxba;hoxbb cluster mutants at 30 hpf (Figure 1H-M). The absence of pectoral fins in zebrafish hoxba;hoxbb cluster mutants sharply contrasts with the results of a previous study, which showed that mice lacking all HoxB genes except for Hoxb13 of the HoxB cluster did not exhibit apparent abnormalities in their forelimbs (Medina-Martinez et al., 2000). Our genetic results suggest that the zebrafish hoxba and hoxbb clusters cooperatively play a crucial role in the pectoral fin formation.

Lack of pectoral fins in hoxba;hoxbb cluster-deleted mutants.

(A-G) Dorsal views of live zebrafish larvae at 3 dpf, obtained from intercrosses between hoxba;hoxbb cluster-deficient hemizygous fish. Arrowheads indicate the positions of the pectoral fins. (H-M) Expression patterns of tbx5a in the pectoral fin bud (arrowhead) at 30 hpf. Dorsal views are shown, and the genotype of each specimen was determined. For each genotype, reproducibility was confirmed with at least three different specimens. Genotyping revealed that all the embryos lacking pectoral fins (n=15) were hoxba;hoxbb double homozygotes. Scale bars are 100 μm.

Lack of pectoral fins is specific to zebrafish hoxba;hoxbb cluster-deleted embryos

Multiple genetic studies in mice have demonstrated the functional redundancy of Hox genes across the four Hox clusters (Horan et al., 1995; McIntyre et al., 2007; van den Akker et al., 2001; Wellik and Capecchi, 2003; Wellik et al., 2002). To investigate whether the absence of pectoral fins is specific to hoxba;hoxbb cluster mutants, we created double-hox cluster mutants by crossing hoxba cluster mutants with six other hox cluster mutants in zebrafish. The combinatorial mutations of hoxba cluster with five other hox clusters, excluding hoxbb cluster, did not enhance the reduced expression of tbx5a (Figure 2A-H). Furthermore, we examined combined mutants of hox clusters, excluding hoxba and hoxbb. In hoxca-/-;hoxcb-/- cluster mutants, where the functions of HoxC genes are completely absent, the expression patterns of tbx5a in fin buds were indistinguishable from those of wild-type embryos (Figure 2I). Additionally, our previous study showed that tbx5a expression is unaffected in hoxaa-/-;hoxab-/-;hoxda-/-cluster mutants (Ishizaka et al., 2024), which lack all HoxA and HoxD-related genes. The results of our deletion mutations of zebrafish hox clusters in various combinations emphasize that the diminished expression of tbx5a in the pectoral fin buds is evident only in hoxba;hoxbb cluster mutants. Taken together, in contrast to the functional redundancy of the four Hox clusters in mice, our genetic analysis indicates that the Hox clusters responsible for the specification of pectoral fin buds are restricted to the HoxB-derived hoxba and hoxbb clusters among the seven hox clusters in zebrafish.

Significantly decreased expression of tbx5a in the pectoral fin buds is specific to hoxba;hoxbb cluster-deleted mutants.

(A-I) Expression patterns of tbx5a in the pectoral fin buds of combinatorial deletion mutants of zebrafish hox clusters. Dorsal views of embryos at 30 hpf are displayed. After capturing images, the genotype of each specimen was determined. For each genotype, reproducibility was confirmed with at least three different specimens. Scale bars are 100 μm.

Transformation of pectoral fin progenitor cells into cardiac cells in hoxba;hoxbb cluster-deficient embryos

To understand the absence of pectoral fin formation in hoxba;hoxbb cluster mutants, we examined the expression patterns of tbx5a during embryogenesis (Figure 3A-F). In wild-type embryos at the 10-somite stage, tbx5a expression appeared as bilateral symmetric stripes in the anterior lateral mesoderm, extending from the posterior midbrain to the second somite (Figure 3A) (Ahn et al., 2002). These tbx5a-positive cells subsequently divided into two groups: the anteriorly migrating cells form the heart primordia, while the posteriorly migrating cells give rise to the future pectoral fin buds (Figure 3C, E). In contrast, we observed that abnormal expression patterns of tbx5a in the lateral mesoderm were evident at the early stages in hoxba;hoxbb cluster mutants. Compared to wild-type embryos, the bilateral stripes of tbx5a expression were shorter along the anterior-posterior axis, and the tbx5a-positive signal was absent from the posterior region where pectoral fin progenitor cells typically emerge (Figure 3B). Although anterior migration toward the presumptive heart occurred, posterior migration toward the fin buds was undetectable due to the lack of pectoral fin progenitor cells in hoxba;hoxbb cluster mutants (Figure 3D, F). These results underscore the early role of the hoxba and hoxbb clusters in providing positional information for pectoral fin buds through the induction of tbx5a expression. In both mice and zebrafish, there is a converse relationship between forelimb/pectoral fin and heart formation in the anterior lateral mesoderm (Keegan et al., 2005; Niederreither et al., 2001; Niederreither et al., 2002; Waxman et al., 2008). Therefore, we also analyzed cardiac development (Figure 3G-J). By examining the expression patterns of nkx2.5, an early cardiac progenitor marker (Chen and Fishman, 1996; Serbedzija et al., 1998), we found that the expression domain of nkx2.5 in hoxba;hoxbb cluster mutants extended more posteriorly compared to wild-type embryos (Figure 3G, H). Furthermore, the region positive for the differentiated cardiac cell-specific myosin light chain (cmlc2) was enlarged in hoxba;hoxbb cluster mutants (Figure 3I, J) (Yelon et al., 1999). These results indicate that hoxba and hoxbb clusters play an essential role in the initial establishment of the pectoral fin field in zebrafish. In their absence, cells that should form the pectoral fins instead differentiate into cardiac cells. The transformation of pectoral fin progenitor cells into cardiac cells observed in zebrafish hoxba;hoxbb cluster mutants is reminiscent of the anterior transformation typical of the loss-of-function phenotypes of Hox genes in Drosophila and mice (Le Mouellic et al., 1992; Lewis, 1978).

Transformation of pectoral fin progenitor cells into cardiac cells in hoxba;hoxbb homozygous embryos.

(A-F) Expression patterns of tbx5a were compared between sibling wild-type and hoxba;hoxbb homozygous embryos during embryogenesis. The range of tbx5a expression in the lateral mesoderm is indicated by a bracket. The progenitor cells of pectoral fins are indicated by an arrowhead. (G, H) Expression patterns of nkx2.5 were compared between sibling and hoxba;hoxbb mutants. Enlarged views are shown in (G’, H’). The different regions of nkx2.5 expression between wild-type and mutants are indicated by brackets. (I, J) Expression patterns of cmlc2 are shown. All images were captured from the dorsal side. For each stage, reproducibility was confirmed with at least three different specimens. Scale bars indicate 100 μm.

hoxba and hoxbb clusters are indispensable for the retinoic acid-mediated induction of tbx5a expression in the fin buds

Retinoic acid (RA) is a well-known upstream regulator of Hox genes in vertebrates (Boncinelli et al., 1991; Langston and Gudas, 1994; Marshall et al., 1996). Loss of function of retinaldehyde dehydrogenase 2 (raldh2), an essential regulator in RA synthesis, leads to the absence of pectoral fins in zebrafish (Begemann et al., 2001; Grandel et al., 2002), a phenotype similar to that observed in hoxba;hoxbb cluster mutants. Previous studies have shown that exogenous RA treatments can rescue pectoral fin formation in raldh2-/- embryos (Gibert et al., 2006; Grandel et al., 2002). It has been suggested that RA directly or indirectly induces tbx5a expression in zebrafish fin buds (Gibert et al., 2006; Grandel and Brand, 2011; Neto et al., 2012). The phenotypic similarities between hoxba;hoxbb cluster mutants and raldh2 mutants prompted us to investigate whether RA exposure could also rescue the absence of pectoral fins in hoxba;hoxbb cluster mutants. To explore this, we introduced a frameshift mutation in raldh2 using CRISPR-Cas9, confirming the phenotypic recapitulation of previously described raldh2-/- mutants (figure-supplement 1). Consistent with prior results (Gibert et al., 2006; Grandel et al., 2002), RA treatment in our raldh2 mutants rescued the expression of tbx5a in pectoral fins (Figure 4A-D). In contrast, RA did not induce tbx5a expression in the fin buds of hoxba;hoxbb cluster mutants (Figure 4E-H), and no trace of pectoral fin formation was observed in the treated embryos even at 5 dpf. Furthermore, the endogenous expression patterns of raldh2 in hoxba;hoxbb cluster mutants seem indistinguishable from those in wild-type embryos (Figure 4I, J), suggesting that the response to RA is absent in the fin buds of hoxba;hoxbb cluster mutants, while RA synthesis appears intact. Although a model proposing that RA directly induces tbx5a expression has been suggested (Gibert et al., 2006; Grandel and Brand, 2011; Neto et al., 2012), our results favor the model that RA-mediated induction of tbx5a in the fin buds occurs through hoxba and hoxbb clusters. Given that hoxba;hoxbb cluster mutants retain the other five intact hox clusters, our findings further support the notion that the hoxba and hoxbb clusters are specifically responsible for the RA-mediated specification of pectoral fins.

Zebrafish hoxba;hoxbb mutants lack a response to RA in fin buds.

(A-D) Exogenous RA exposure in raldh2 mutants can rescue tbx5a expression in fin buds. Arrowheads indicate the positions of the pectoral fin buds. (E-H) RA treatments in hoxba;hoxbb cluster-deleted mutants do not rescue tbx5a expression in fin buds. (I, J) Expression patterns of raldh2 were compared between sibling and hoxba;hoxbb mutants. For each genotype, reproducibility was confirmed with at least three different specimens. All images were captured from the dorsal side. Scale bars represent 100 μm.

Expression patterns of hoxb5a and hoxb5b in pectoral fin buds are dependent on RA

The absence of pectoral fins, commonly observed in hoxba;hoxbb cluster mutants and raldh2 mutants, suggests that candidate hox genes may be downregulated in raldh2 mutants. To investigate this, we performed RNA-seq analysis to compare the expression profiles between sibling and raldh2 homozygous embryos. As expected, we found that most of the 3’-located hox genes are downregulated in raldh2-/- embryos (Figure 5A, B, figure supplement 2). Notably, the levels of hoxb5b transcripts were significantly reduced, along with those of hoxb5a among hoxba genes, in raldh2 mutants. Previous studies have shown that zebrafish hoxb5a and hoxb5b are expressed in the lateral plate mesoderm corresponding to presumptive pectoral fin buds (Waxman et al., 2008). Additionally, knockout mice for Hoxb5, which is homologous to zebrafish hoxb5a and hoxb5b, only exhibit altered positioning of forelimbs (Rancourt et al., 1995), among various other Hox knockouts that have been generated. Therefore, we examined whether RA regulates the expression of hoxb5a and hoxb5b in fin buds. In wild-type embryos, hoxb5a expression was observed in the neural tube, somites, and lateral plate mesoderm (Figure 5C). However, in raldh2 mutants, hoxb5a expression was significantly reduced, notably in the lateral plate mesoderm, which differentiates into the fin buds, where it was reduced to the point of being undetectable (Figure 5D). Since RA treatments in raldh2 mutants can rescue pectoral fin development, we found that hoxb5a expression in raldh2 mutants is also rescued by RA exposure (Figure 5C-E). Regarding the expression of hoxb5b, which diverged from the same ancestral gene as hoxb5a, RA-dependent expression was also confirmed, mirroring the observations made with hoxb5a (Figure 5F-H). These results suggest that both hoxb5a and hoxb5b are involved in the formation of the pectoral fins in zebrafish.

The expression patterns of hox5a and hoxb5b are regulated by RA.

(A) Schematic representation of the 49 hox genes organized into seven hox clusters in zebrafish. hox genes in hoxba and hoxbb clusters are highlighted with orange. (B)Expression profiles of the 49 hox genes in wild-type and raldh2-/- embryos at the 20-somite stage, analyzed by RNA-seq. The average FPKM of each hox gene was compared between sibling and raldh2 mutants. The absence of specific hox genes is indicated by a slash. (C-H) Expression patterns of hoxb5a and hoxb5b in sibling wild-type, raldh2-/-, and RA-treated raldh2-/- embryos at the 10-somite stage. Arrowheads indicate the presumptive positions of pectoral fin buds. For each genotype, reproducibility was confirmed with at least three different specimens. All images are captured from the dorsal side. Scale bars represent 100 μm.

Genetic screening for hox genes responsible for the specification of pectoral fin buds in zebrafish

To understand the molecular mechanisms underlying pectoral fin specification, we aimed to identify the hox genes essential for pectoral fin formation through a genetic approach. By introducing frameshift mutations into the relevant hox genes, we anticipated that the resulting mutants would replicate the absence of pectoral fins observed in hoxba;hoxbb mutants. Since hoxbb cluster contains fewer hox genes than hoxba cluster (Figure 5A), we focused our investigation on hoxbb cluster. Although hoxb5b is expressed in the presumptive fin buds (Figure 5F), other hoxbb genes may also be involved. Therefore, we simultaneously injected three gRNAs targeting hoxb5b, hoxb6b, and hoxb8b into hoxba hemizygous mutants, expecting various germline mutations in these hoxbb genes in the injected fish (Figure 6A). Subsequently, F1 fish with a hemizygous deletion of hoxba cluster and mosaic mutations were crossed with hoxba;hoxbb hemizygous mutants. Among the progeny, several embryos lacking pectoral fins were identified. Following genotyping and DNA sequencing of the target hoxbb genes, we found that all embryos without pectoral fins possessed frameshift mutations in hoxb5b, with hoxba-/-;hoxbb+/- being common among them (Figure 6B-E, table-supplement 1, figure-supplement 3). Similarly, we injected each crRNA for hoxb6b and hoxb8b into hoxba+/- mutants and confirmed that neither frameshift mutation in hoxb6b nor hoxb8b in hoxba-/-;hoxbb+/- led to the loss of pectoral fins (Figure 6F, G, figure-supplement 3). These results suggest that the frameshift mutation of hoxb5b alone can compensate for the deletion of hoxbb cluster, recapitulating the absence of pectoral fins in hoxba;hoxbb cluster mutants.

Screening for hox genes responsible for the specification of zebrafish pectoral fins.

(A) Schematic representation of the genetic screening for hox genes involved in the specification of pectoral fins in zebrafish. (B-G) Dorsal views of live mutant larvae at 3 dpf, were obtained during the screening. After capturing images, genotyping and DNA sequencing in target hoxb genes were conducted. For the mutants illustrated in (C-G), hoxb genes are shown with frameshift mutations introduced. Detailed information is provided in Supplementary Table S1. (H-J) Dorsal views of hoxba-/-;hoxb5b-/- and hoxba-/-;hoxb5b-/-;hoxb6b-/- larvae at 3 dpf. (K-M) Expression patterns of tbx5a in hoxba-/-;hoxb5b-/- and hoxba-/-;hoxb5b-/-;hoxb6b-/- mutants at 30 hpf. Arrowheads indicate the presumptive positions of pectoral fin buds. All images are captured from the dorsal side. Scale bars represent 100 μm.

However, we encountered a problem: hoxba-/-;hoxb5b-/- embryos, which have a homozygous frameshift mutation in hoxb5b and a homozygous deletion of hoxba cluster, exhibited severely truncated but visibly present pectoral fins (Figure 6H, I). After analyzing hundreds of embryos from several crossings, we did not find any hoxba-/-;hoxb5b-/- embryos without pectoral fins. We also confirmed that the expression of tbx5a was detectable in the fin buds of hoxba-/-;hoxb5b-/- embryos (Figure 6K, L), in contrast to its absence in hoxba;hoxbb cluster mutants (Figure 1M). To examine the potential contributions of other hoxbb genes, we observed the phenotypes of hoxba-/-;hoxb5b-/-; hoxb6b-/- embryos; however, these mutants did not exacerbate the abnormalities and did not also exhibit the absence of pectoral fins (Figure 6J, M).

The phenotype of hoxba-/-;hoxb5b-/- embryos indicates that the frameshift mutation in hoxb5b cannot fully compensate for the complete deletion of hoxbb cluster, suggesting that other mechanisms may be involved. A novel genetic compensation mechanism known as transcriptional adaptation has been identified: mRNA containing a premature termination codon (PTC) can induce increased expression of structurally related genes, thereby compensating for the function of mutated genes (El-Brolosy et al., 2019; Rossi et al., 2015). To avoid producing transcripts with PTC (Sztal and Stainier, 2020), we generated a full locus-deleted allele of hoxb5b using CRISPR-Cas9 with two crRNAs targeting both ends of the target locus (figure-supplement 4). In contrast to the phenotype of hoxba-/-;hoxb5b-/-, we found that hoxba-/-;hoxb5b de/dell mutants lack pectoral fins and exhibit a significant reduction in tbx5a expression when intercrossing hoxba+/-;hoxb5b+/del fish (Figure 7A, B, D, E). However, the occurrence rate of embryos lacking pectoral fins (n=3/120; 2.5 %) was lower than predicted based on Mendelian genetics (1/16; 6.3 %). Other hoxba-/-;hoxb5bdel/del embryos did exhibit shortened pectoral fins (Figure 7C). Additionally, concerning hoxba cluster, we also generated frameshift-induced hoxb5a-/-;hoxb5b-/- and a full locus-deleted hoxb5adel/del;hoxb5bdel/delembryos (figure-supplement 4). hoxb5a-/-;hoxb5b-/-larvae consistently displayed normal pectoral fins (Figure 7F). In contrast, hoxb5adel/del;hoxb5bdel/del mutants exhibited a slightly more pronounced phenotype of shortened pectoral fins, but no instances of missing pectoral fins were identified (Figure 7G). Furthermore, due to the functional redundancy observed between hoxb4a and hoxb5a observed in zebrafish vertebral patterning and the RA-dependent expression of hoxb4a (Grandel et al., 2002; Maeno et al., 2024), as well as its expression in fin buds (Figure 7H), we generated frameshift-induced hoxb4a-/-;hoxb5a-/-;hoxb5b-/- embryos and an allele lacking the entire genomic region encompassing hoxb4a and hoxb5a (figure-supplement 4). Although hoxb4a-/-;hoxb5a-/-;hoxb5b-/- embryos did not lack pectoral fins (Figure 7I), some hoxb4a-b5adel/del;hoxb5b del/del embryos did lack pectoral fins, although at low penetrance (Figure 7J, n=3/397; 0.7%). The majority of hoxb4a-b5adel/del;hoxb5b del/del embryos displayed truncated pectoral fins (Figure 7K). The lack of pectoral fins was not detected in hoxb4a-b5adel/del and hoxb4adel/del;hoxb5bdel/del mutants (Figure 7L, M, (figure-supplement 4). These results suggest that hoxb4a, hoxb5a, and hoxb5b may cooperatively contribute to anterior-posterior positioning of pectoral fin buds in zebrafish. Furthermore, our findings reveal that transcriptional adaptation may partially play a role in genetic compensation for frameshift-induced hox mutations during pectoral fin formation, while also suggesting the potential involvement of other unknown mechanisms or additional genomic regions in the initial induction of zebrafish pectoral fin buds.

hoxb4a-b5adel/del;hoxb5bdel/del larvae partially recapitulate the absence of the pectoral fins.

(A-C) Dorsal views of zebrafish hoxba-/-;hoxb5bdel/del larvae at 3 dpf. (D, E) Expression patterns of tbx5a in sibling wild-type and hoxba-/-;hoxb5bdel/del (n=4) at 30 hpf. Arrowheads indicate the presumptive positions of pectoral fin buds. (F, G) Dorsal view of frameshift-induced hoxb5a-/-;hoxb5b-/- and hoxb5a del/del;hoxb5bdel/dellarvae. (H) Expression patterns of zebrafish hoxb4a at the 10-somite stage. (I-K) Dorsal view of frameshift-induced hoxb4a-/-;hoxb5a-/-;hoxb5b-/- and hoxb4a-b5a del/del;hoxb5bdel/del larvae. All images are captured from the dorsal side. Arrowheads indicate the presumptive positions of pectoral fin buds. Scale bars represent 100 μm.

Discussion

Since the establishment of gene targeting (Mansour et al., 1988), numerous Hox knockout mice have been generated. However, neither single nor compound Hox mutants, nor Hox overexpression approaches, have been reported to cause significant defects in limb positioning (Jurberg et al., 2013; Moreau et al., 2019). Consequently, the role of Hox genes in the positioning of vertebrate paired appendages remains unclear due to a lack of genetic evidence. In this study, we provide genetic evidence using zebrafish, demonstrating that specific combinatorial deletions of hoxba and hoxbb clusters result in the absence of pectoral fins and a lack of induction of tbx5a expression in the pectoral fin field. Based on our results, we propose a model in which Hox expressions in the lateral plate mesoderm, regulated by RA synthesized in the paraxial mesoderm, provide positional information and induce tbx5a expression in the presumptive fin buds. Our genetic results shed light on longstanding fundamental questions regarding the molecular mechanisms by which the positioning of paired appendages is established in vertebrates.

Zebrafish hoxba and hoxbb clusters, possibly hoxb4a, hoxb5a, and hoxb5b, cooperatively define the pectoral fin field in the lateral plate mesoderm

We showed that hoxba;hoxba mutants exhibit a lack of pectoral fins. The abnormal expression of tbx5a in the lateral plate mesoderm, where pectoral fin precursor cells arise, was evident from its initial expression in the mutants. We presume that hox genes in hoxba and hoxbb clusters, induced by RA, define the region along the anterior-posterior axis where pectoral fin formation can occur and promote the expression of tbx5a in the lateral plate mesoderm. Importantly, tbx5a expression in the fin bud was not induced even by RA exposure in hoxba;hoxbb mutants (Figure 4E-H). In zebrafish, it has been suggested that RA synthesized in the paraxial mesoderm acts directly on the lateral plate mesoderm, or that RA induces tbx5a expression via wnt2b in the intermediate mesoderm (Gibert et al., 2006; Grandel and Brand, 2011; Neto et al., 2012). However, our results suggest that the main pathway integrates these signaling cascades through hoxba and hoxbb clusters to induce tbx5a expression. Through genetic studies, despite low penetrance, we showed that hoxb4a, hoxb5a, and hoxb5b play a significant role in the induction of pectoral fin buds. The homologous Hoxb4 and Hoxb5 proteins in mice have been shown to bind directly to the enhancer of Tbx5 (Minguillon et al., 2012). Furthermore, it has been demonstrated that the Tbx5 expression in the forelimb field is restricted by the posterior Hox genes, preventing posterior expansion in mice and chickens (Moreau et al., 2019; Nishimoto et al., 2014). However, our several combinatorial deletions of zebrafish hox clusters did not exhibit any abnormalities in which the expression of tbx5a expanded to the posterior. Therefore, in zebrafish, we propose that hoxb4a, hoxb5a, and hoxb5b, which are induced by RA in the lateral plate mesoderm, simply upregulate the expression of tbx5a.

Partial recapitulation of genomic locus-deleted hox mutants suggests the other unidentified mechanisms

In this study, we observed that hox mutants with deletions of genomic loci exhibited more pronounced pectoral fin phenotypes than frameshift-induced hox mutants. The frameshift mutants used in this study are likely to be loss-of-function alleles, as we demonstrated several defects in vertebral patterning in these mutants (Maeno et al., 2024). These results suggest that transcriptional adaptation may occur during the formation of pectoral fins in frameshift-induced hox mutants, potentially compensating for the loss of function. In our previous studies, which focused on dorsal and anal fin patterning, vertebral patterning, and swim bladder formation, we demonstrated that the phenotypes observed in zebrafish hox cluster-deleted mutants could essentially be replicated by introducing frameshift mutations into hox genes within the cluster (Adachi et al., 2024; Maeno et al., 2024; Satoh et al., 2024; Yamada et al., 2021). Therefore, creating a mutant with a deleted locus that does not induce transcriptional adaptation was unnecessary. One factor that may explain the differences in results between this study and our prior findings is that even a small amount of hox gene products essential for pectoral fin positioning may be sufficient to induce fin buds, as shown in Figure 1, potentially leading to the manifestation of gene compensation through transcriptional adaptation in frameshift-induced mutants. Alternatively, transcriptional adaptation may function differently depending on various developmental processes or stages. Interestingly, all the aforementioned phenotypes reproducible by frameshift mutations were observed at later developmental stages than pectoral fin formation. Additionally, in hox locus-deleted mutants, only some hoxba-/-;hoxb5b de/dell or hoxb4a-b5adel/del;hoxb5b del/del embryos were able to reproduce the loss of the pectoral fin, suggesting the presence of other compensatory mechanisms. The organized expression patterns of Hox genes are thought to be regulated by complex regulatory controls, with their expression regulated according to the genomic order within the cluster. In the absence of Hox function, neighboring intact Hox genes could compensate for the loss of function. It is also possible that unidentified non-coding regions in hoxba and hoxbb clusters are involved in pectoral fin positioning. These issues should be clarified in future studies.

Hox genes may have contributed to the acquisition of pectoral fins during vertebrate evolution

According to fossil records from the Cambrian period, the most primitive vertebrates did not possess paired fins (Janvier, 1996; Shubin et al., 1997). It is believed that, during subsequent evolution, vertebrates acquired primitive pectoral and pelvic fins, which served as evolutionary precursors to tetrapod limbs, before diverging into ray-finned and lobe-finned fishes. Our results align well with this evolutionary scenario. Hox genes, which are conserved in bilaterian animals, are thought to have been present in primitive vertebrates without paired fins. The early fossils of vertebrates with pectoral fins were discovered in Ordovician and Silurian jawless fish (Janvier, 1996; Shubin et al., 1997). Since it is estimated that two rounds of whole-genome duplications occurred early in the Cambrian period (Dehal and Boore, 2005), the first vertebrate with pectoral fins likely possessed four sets of Hox clusters, similar to those found in modern tetrapods. Hox genes have long been fundamental tools for animal development, even before vertebrates acquired paired fins. The acquisition of new functions in specific Hox genes may have contributed to the innovation of novel morphological features. Our findings that hoxba;hoxbb mutants completely lack pectoral fins suggest that molecular evolution within the ancestral HoxB cluster may have contributed to the development of pectoral fins in vertebrates. Although the specific molecular modifications remain unclear, one possible model suggests that HoxB gene(s) acquired a new capacity to induce the expression of Tbx5 in the lateral plate mesoderm. This may have played a role in establishing the RA-HoxB-Tbx5 gene cascade, which is essential for pectoral fin formation. Notably, the ancestral Tbx4/5 in ascidians, which do not have paired fins, already possesses the ability to rescue the loss of forelimbs in mice (Minguillon et al., 2009). Consistent with our idea, a model has been proposed in which the induction of Tbx5 in the lateral plate mesoderm may have contributed to the acquisition of pectoral fins.

Since hoxba;hoxbb mutants are embryonic lethal around 5 dpf, we cannot observe the effects of these clusters on the pelvic fins, which develop around one month after fertilization in zebrafish (Parichy et al., 2009). However, based on the composition of zebrafish hox genes and their expression patterns in hoxba and hoxbb clusters, we presume that these clusters have functions specific to the positioning of pectoral fins. The presence of hox genes responsible for the position of pectoral fins implies that there may be other hox genes that determine the position of pelvic fins and possibly contribute to their acquisition. This remains a critical area for future research.

The positioning of pectoral fins is restricted to specific hox clusters in zebrafish, in contrast to the functional overlap seen in mice

Taking advantage of our availability of seven individual hox cluster mutants in zebrafish, we created mutants with multiple cluster deletions. Through genetic analysis, we demonstrated that hox genes responsible for positioning pectoral fin buds are restricted to hoxba and hoxbb clusters. In contrast, analyses of knockout mice reveal significant functional overlap among the four Hox clusters. Mice lacking all HoxB genes except for Hoxb13 have been generated, yet there are no reports of a phenotype characterized by absent forelimbs (Medina-Martinez et al., 2000). In Hoxb5 knockout mice, the position of the forelimbs shifts rostrally by one vertebra, and forelimbs are not absent in Hoxa5;Hoxb5;Hoxc5 triple knockout mice, which lack all Hox genes of the paralog 5 group (Xu et al., 2013). In comparison, zebrafish hoxba;hoxbb mutants exhibit a striking phenotype of absent pectoral fins. This phenomenon, where function is confined to specific hox clusters, has also been observed in other developmental systems of teleosts. In both zebrafish and medaka, multiple hox genes responsible for positioning dorsal and anal fins are primarily located in HoxC-related clusters (Adachi et al., 2024). As mentioned earlier, it is believed that paired fins in vertebrates emerged after the quadruplication of the primitive Hox clusters. Therefore, Hox genes that acquired new functions related to pectoral fin formation likely originated from specific Hox clusters among the four, maintaining restricted functions that diverged after the teleost-specific whole-genome duplication. The validity of this assumption requires further verification through a clearer understanding of the overall functions of the 49 hox genes in zebrafish. However, the genetic results thus far suggest that, unlike in mice, the functions of hox genes—especially those related to development uniquely acquired in vertebrates—may be confined to specific hox clusters in zebrafish. While teleosts have gained functional redundancy through subsequent teleost-specific whole-genome duplication, it is postulated that, in mice, which have maintained four Hox clusters for over half a billion years, functional overlap among Hox genes may ensure the robustness of the developmental system.

Materials and methods

Zebrafish

Riken Wild-type (RW) zebrafish, obtained from the National BioResource Project in Japan, were maintained at 27°C with a 14-hour light/10-hour dark cycle. Embryos were collected from natural spawning, and the larvae were raised at 28.5°C. Developmental stages of the embryos and larvae were determined based on hours post-fertilization (hpf), days post-fertilization (dpf), and developmental stage-specific features (Kimmel et al., 1995; Parichy et al., 2009). The alleles of hox cluster-deleted mutants used in this study were previously generated using CRISPR-Cas9 (Yamada et al., 2021). The following frameshift-induced hox mutants were created using CRISPR-Cas9: hoxb5asud125, hoxb5b sud136, hoxb6b sud137, and hoxb4a;hoxb5a sud144 (Maeno et al., 2024). All experiments involving live zebrafish were approved by the Committee for Animal Care and Use of Saitama University.

Generation of mutants by the CRISPR-Cas9 system

All the mutants used in this study were generated using the Alt-R CRISPR-Cas9 system (Integrated DNA Technologies). To create frameshift-induced mutants, gene-specific crRNAs were incubated with common tracrRNA, followed by Cas9 nuclease. For the generation of mutants that delete large genomic regions, two crRNAs targeting both ends of the regions were incubated with common tracrRNA and Cas9 nuclease. Approximately one nanoliter of the crRNA:tracrRNA-Cas9 RNA-protein complex was injected into fertilized embryos, which were subsequently raised to juvenile fish. Candidate founder fish were selected using a heteroduplex mobility shift assay for frameshift mutants (Ota et al., 2013) and by amplifying genomic deletions with specific primers. After sexual maturation, candidate founder fish were mated with wild-type fish to produce heterozygous F1 offspring. Among the F1 offspring, mutant fish carrying the same mutation were identified through PCR-based genotyping followed by DNA sequencing. The target-specific sequences of the crRNAs used in this study are listed (table-supplement 2). The frozen sperms from the mutants used in this study have been deposited in the National BioResource Project Zebrafish in Japan (https://shigen.nig.ac.jp/zebra/) and are available upon request.

Genotyping of mutants

For the phenotypic analysis of embryos and the maintenance of mutant fish lines, PCR-based genotyping was performed. Briefly, genomic DNA was extracted from the analyzed embryos or the partially dissected caudal fins of anesthetized larvae or fish using the NaOH method (Meeker et al., 2007), which served as the template for PCR. The sequences of the primers used for genotyping frameshift-induced mutants are listed in the supplementary Table 3. For genotyping deletion mutants of hox genes, PCR was conducted using a combination of three primers, with their sequences also provided (table-supplement 3). Genotyping of hox cluster-deleted mutants was performed using PCR as previously described (Yamada et al., 2021). After the reactions, the PCR products were separated by electrophoresis. Based on differences in the lengths of PCR products derived from wild-type and mutated alleles, either a 2 % agarose gel in 0.5x TBE buffer, a 15 % polyacrylamide gel in 0.5x TBE buffer, or direct DNA sequencing was utilized to determine the genotype.

Whole-mount in situ hybridization

Whole-mount in situ hybridization was carried out as described (Thisse and Thisse, 2014). After the staining, the embryos mounted in 70-80 % glycerol were captured under the stereomicroscope (Leica M205 FA) with a digital camera (Leica DFC350F). After taking images, PCR-based genotyping was carried out as described above.

RNA-seq analysis

For RNA-seq analysis, embryos were obtained by the intercrosses between raldh2 heterozygous fish. At the 18-20-somite stages, sibling and homozygous embryos were separated based on the phenotype, and 20 embryos per tube were collected. Then, total RNA was extracted by ISOGENE (Nippon Gene) and followed by DNase I treatment (Takara). After the quality check of the isolated RNA, RNA-seq analyses (n=2 for each) were carried out by Genewiz, Azenta Life Science.

Treatment of retinoic acid in the embryos

Treatment of retinoic acid on the embryos was performed as previously described, with minor modifications (Gibert et al., 2006; Grandel et al., 2002). Briefly, embryos were obtained from intercrosses between raldh2 heterozygous mutants. At the 70 % epiboly stage, the embryos were soaked in a solution containing DMSO or 10-8M RA and incubated at 28.5°C. Embryos were fixed at 30 hpf for analysis of tbx5a expression and at the 10-somite stage for examination of hox gene expression.

eLife Assessment

This important study advances our understanding of vertebrate forelimb development, specifically the contribution of Hox genes to zebrafish pectoral fin formation. While there are reservations about some of the descriptions and interpretations of the data, the results are mostly convincing. The authors have employed a robust and extensive genetic approach to tackle a key and unresolved question. The findings will be of broad interest to developmental and evolutionary biologists.

Reviewer #1 (Public review):

Summary:

The authors have used gene deletion approaches in zebrafish to investigate the function of genes of the hox clusters in pectoral fin "positioning" (but perhaps more accurately pectoral fin "formation").

Strengths:

The authors have employed a robust and extensive genetic approach to tackle an important and unresolved question.

The results are largely presented in a very clear way.

Weaknesses:

The Abstract suggests that no genetic evidence exists in model organisms for a role of Hox genes in limb positioning. There are, however, several examples in mouse and other models (both classical genetic and other) providing evidence for a role of Hox genes in limb position, which is elaborated on in the Introduction.

It would perhaps be more accurate to state that several lines of evidence in a range of model organisms (including the mouse) support a role for Hox genes in limb positioning. The author's work is not weakened by a more inclusive introduction that cites the current literature more comprehensively.

It would be helpful for the authors to make a clear distinction between "positioning" of the limb/fin and whether a limb/fin "forms" at all, independent of the relative position of this event along the body axis.

Discussion of why the zebrafish is sensitive to Hoxb loss with reference to the fin, but mouse Hoxb mutants do make a limb?

Is this down to exclusive expression of Hoxbs in the zebrafish pectoral fin forming region rather than a specific functional role of the protein? This is important as it has implications for the interpretation of results throughout the paper and could explain some apparently conflicting results.

Why is Hoxba more potent than Hoxbb? Is this because Hoxba has Hox4/5 present, while Hoxbb has only Hoxb5? Hoxba locus has retained many more Hox genes in cluster than hoxbb; therefore, one might expect to see greater redundancy in this locus).

Deletion of either Hoxa or Hoxd in the background of the Hoxba mutant does have some effect. Is this a reflection of protein function or expression dynamics of Hoxa/Hoxd genes?

Can we really be confident that there is a "transformation of pectoral fin progenitor cells into cardiac cells"?

The failure to repress Nkx2.5 in the posterior (pelvic fin) domain is clear, but have these cells actually acquired cardiac identity? They would be expected to express Tbx5a (or b) as cardiac precursors, but this domain does not broaden. There is no apparent expansion of the heart (field)/domain or progenitors beyond the 16 somite stage. The claimed "migration" of heart precursors in the mutant is not clear. The heart/cardiac domain that does form in the mutant is not clearly expanded in the mutant. The domain of cmlc2 looks abnormal in the mutant, but I am not convinced it is "enlarged" as claimed by the authors. The authors have not convincingly shown that "the cells that should form the pectoral fin instead differentiate into cardiac cells."

The only clear conclusion is the loss of pectoral fin-forming cells rather than these fin-forming cells being "transformed" into a new identity. It would be interesting to know what has happened to the cells of the pectoral fin-forming region in these double mutants.

It is not clear what the authors mean by a "converse" relationship between forelimb/pectoral fin and heart formation. The embryological relationship between these two populations is distinct in amniotes.

The authors show convincing data that RA cannot induce Tbx5a in the absence of Hob clusters, but I am not convinced by the interpretation of this result. The results shown would still be consistent with RA acting directly upstream of tbx5a, but merely that RA acts in concert with hox genes to activate tbx5a. In the absence of one or the other, Tbx5a would not be expressed. It is not necessary that RA and hoxbs act exclusively in a linear manner (i.e., RA regulates hoxb that in turn regulates tbx5a).

The authors have carried out a functional test for the function of hoxb6 and hoxb8 in the hemizygous hoxb mutant background. What is lacking is any expression analysis to demonstrate whether Hoxb6b or Hoxb8b are even expressed in the appropriate pectoral fin territory to be able to contribute to pectoral fin development, either in this assay or in normal pectoral fin development.

(The term "compensate" used in this section is confusing/misleading.)

The authors' confounding results described in Figures 6-7 are consistent with the challenges faced in other model organisms in trying to explore the function of genes in the hox cluster and the known redundancy that exists across paralogous groups and across individual clusters.

Given the experimental challenges in deciphering the actual functions of individual or groups of hox genes, a discussion of the normal expression pattern of individual and groups of hox genes (and how this may change in different mutant backgrounds) could be helpful to make conclusions about likely normal function of these genes and compensation/redundancy in different mutant scenarios.

Reviewer #2 (Public review):

Summary:

The authors of this manuscript performed a fascinating set of zebrafish mutant analyses on hox cluster deletion and pinpointed the cause of the pectoral fin loss in one combinatorial hox cluster mutant of Hoxba and Hoxbb.

Strengths:

The study is based on a variety of existing experimental tools that enabled the authors' past construction of hox cluster mutants, and is well-designed. The manuscript is well written to report the authors' findings on the mechanism that positions the pectoral fin.

Weaknesses:

The study does not focus on the other hox clusters other than ba and bb, and is confined to the use of zebrafish, as well as the comparison with existing reports from mouse experiments.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

The authors have used gene deletion approaches in zebrafish to investigate the function of genes of the hox clusters in pectoral fin "positioning" (but perhaps more accurately pectoral fin "formation").

Strengths:

The authors have employed a robust and extensive genetic approach to tackle an important and unresolved question. The results are largely presented in a very clear way.

We thank the reviewer for the positive summary and for recognizing the strengths of our genetic approach and presentation.

Weaknesses:

The Abstract suggests that no genetic evidence exists in model organisms for a role of Hox genes in limb positioning. There are, however, several examples in mouse and other models (both classical genetic and other) providing evidence for a role of Hox genes in limb position, which is elaborated on in the Introduction.

It would perhaps be more accurate to state that several lines of evidence in a range of model organisms (including the mouse) support a role for Hox genes in limb positioning. The author's work is not weakened by a more inclusive introduction that cites the current literature more comprehensively.

Thank you for this constructive comment. We agree that our Abstract implied an absence of genetic evidence across model organisms and could be misleading. We have revised the Abstract to acknowledge that multiple lines of evidence—including classical and molecular studies in mouse and other models—support a role for Hox genes in limb/fin positioning. We have also expanded the Introduction to cite this literature more comprehensively. These changes clarify the current state of knowledge while preserving the novelty of our zebrafish genetic findings.

It would be helpful for the authors to make a clear distinction between "positioning" of the limb/fin and whether a limb/fin "forms" at all, independent of the relative position of this event along the body axis.

We thank the reviewer for pointing this out. In the revised manuscript, we now make a distinction between these two aspects: we describe “positioning” as being specified by the expression domains of Hox genes along the anterior–posterior axis, while the “formation” of pectoral fins reflects the functional requirement of Hox genes to induce tbx5a expression and thereby initiate fin development. We have clarified this distinction in the text to better separate these related but distinct roles of Hox genes.

Discussion of why the zebrafish is sensitive to Hoxb loss with reference to the fin, but mouse Hoxb mutants do make a limb?

We thank the reviewer for this important comment. Our interpretation is that paired fins first appeared in vertebrates that already possessed four Hox clusters. It is likely that novel functions related to pectoral fin positioning emerged within the HoxB cluster at that time, contributing to the origin of pectral fins. In zebrafish, we found that these functions remain largely restricted to the hoxba and hoxbb clusters, such that loss of both results in complete absence of pectoral fins. In contrast, mice exhibit a high degree of functional redundancy across Hox clusters. For example, deletion of all HoxB genes except Hoxb13 does not result in forelimb loss (Medina-Martinez et al., 2000), and forelimbs are still present in Hoxa5;Hoxb5;Hoxc5 triple knockouts (Xu et al., 2013). Thus, although we cannot fully explain why HoxB cluster deletions alone do not abolish forelimb formation in mice, it is plausible that overlapping functions from other Hox clusters compensate for the loss of HoxB genes, consistent with the general robustness of the mammalian Hox system. We have revised the Discussion to clarify this point.

Is this down to exclusive expression of Hoxbs in the zebrafish pectoral fin forming region rather than a specific functional role of the protein? This is important as it has implications for the interpretation of results throughout the paper and could explain some apparently conflicting results.

We thank the reviewer for this insightful comment. To address this point, we newly analyzed the expression patterns of PG4–8 genes in the hoxba and hoxbb clusters. Our in situ hybridization results revealed that only hoxb4a, hoxb5a, and hoxb5b are detectably expressed in the pectoral fin buds (Figure 5C, 5E, Figure 7M-R). While we cannot completely exclude the possibility of functional differences among Hox proteins, our data strongly suggest that the loss of pectoral fins in hoxba;hoxbb cluster mutants is primarily due to the expression domains of these specific Hox genes in the fin-forming region, rather than to unique biochemical properties of the proteins. We have added these new data as a figure in the revised manuscript (Figure 7M-R) and clarified this point in the text (lines 312-316).

Why is Hoxba more potent than Hoxbb? Is this because Hoxba has Hox4/5 present, while Hoxbb has only Hoxb5? Hoxba locus has retained many more Hox genes in cluster than hoxbb; therefore, one might expect to see greater redundancy in this locus).

We thank the reviewer for raising this important point. At present, we do not know the precise reason why hoxba appears more potent than hoxbb. The possibility raised by the reviewer—that differences in retained gene content (e.g., Hox4/5 in hoxba versus only Hoxb5 in hoxbb) may underlie this discrepancy—is certainly plausible. However, our previous study on the formation of dorsal and anal fins showed a similar situation: although PG11–13 Hox genes are present in both hoxca and hoxcb clusters, deletion of hox genes in hoxca cluster had a more pronounced effect on median fin development (Adachi et al., 2024). This suggests that, following the teleost-specific whole-genome duplication, duplicated Hox clusters are not functionally equivalent, and asymmetric retention or deployment of functions may occur. The mechanistic basis of such bias remains unclear and warrants further investigation.

Deletion of either Hoxa or Hoxd in the background of the Hoxba mutant does have some effect. Is this a reflection of protein function or expression dynamics of Hoxa/Hoxd genes?

We appreciate the reviewer’s comment and the opportunity to clarify this point. In Figure 2, we compared several double mutants with the hoxba single mutant. Among thesm, only the hoxba;hoxbb mutant exhibited a complete loss of tbx5a expression, whereas other combinations did not differ substantially from the hoxba mutant alone. Therefore, we consider that additional deletions such as hoxaa, hoxab, and hoxda do not have a strong effect beyond the hoxba deletion itself, and it is unlikely that Hoxa or Hoxd proteins functionally compensate for Hoxba in regulating tbx5a expression. Consistent with this interpretation, in our previous study we did not detect abnormalities in tbx5a expression in the hoxaa;hoxab;hoxda triple mutant (Ishizaka et al., 2024). Taken together, these observations support our view that the hoxba and hoxbb clusters are specifically required for the induction of tbx5a in the pectoral fin field.

Can we really be confident that there is a "transformation of pectoral fin progenitor cells into cardiac cells"?

The failure to repress Nkx2.5 in the posterior (pelvic fin) domain is clear, but have these cells actually acquired cardiac identity? They would be expected to express Tbx5a (or b) as cardiac precursors, but this domain does not broaden. There is no apparent expansion of the heart (field)/domain or progenitors beyond the 16 somite stage. The claimed "migration" of heart precursors in the mutant is not clear. The heart/cardiac domain that does form in the mutant is not clearly expanded in the mutant. The domain of cmlc2 looks abnormal in the mutant, but I am not convinced it is "enlarged" as claimed by the authors. The authors have not convincingly shown that "the cells that should form the pectoral fin instead differentiate into cardiac cells." The only clear conclusion is the loss of pectoral fin-forming cells rather than these fin-forming cells being "transformed" into a new identity. It would be interesting to know what has happened to the cells of the pectoral fin-forming region in these double mutants.

We sincerely thank the reviewer for this important comment. We agree that our data do not yet allow us to conclude with certainty that the presumptive pectoral fin progenitor cells in hoxba;hoxbb cluster mutants are fully “transformed” into cardiac cells. Our intention was to describe the striking posterior expansion of nkx2.5 expression and the altered morphology of the cmlc2-positive cardiac field in the mutants, which suggested a shift in cell fate. However, as the reviewer correctly points out, we did not directly demonstrate that the missing fin progenitors acquire bona fide cardiac identity.

To address this, we have revised the text to clarify that the most robust conclusion from our current dataset is the loss of pectoral fin-forming cells in hoxba;hoxbb cluster mutants. We have softened or removed the claim of “transformation” and instead emphasize that our observations are consistent with an expansion of cardiac marker expression domains into the region where fin progenitors normally arise. We also acknowledge that the cmlc2 domain is abnormal rather than unequivocally enlarged, and have adjusted our wording accordingly.

It is not clear what the authors mean by a "converse" relationship between forelimb/pectoral fin and heart formation. The embryological relationship between these two populations is distinct in amniotes.

We thank the reviewer for pointing this out. Our intention was to highlight the reciprocal balance between pectoral fin and cardiac progenitors in zebrafish. In particular, Waxman et al. (2008) demonstrated that retinoic acid signaling promotes pectoral fin formation while restricting the expansion of cardiac progenitors, thereby illustrating this reciprocal relationship. To avoid confusion, we have revised the text to explicitly state that this applies to zebrafish.

The authors show convincing data that RA cannot induce Tbx5a in the absence of Hob clusters, but I am not convinced by the interpretation of this result. The results shown would still be consistent with RA acting directly upstream of tbx5a, but merely that RA acts in concert with hox genes to activate tbx5a. In the absence of one or the other, Tbx5a would not be expressed. It is not necessary that RA and hoxbs act exclusively in a linear manner (i.e., RA regulates hoxb that in turn regulates tbx5a).

We appreciate the reviewer’s thoughtful comment. We agree that our original wording in the Results section implied a strictly linear model of RA→Hox→tbx5a. In response, we have revised the Results to state only the experimental observation, namely that RA-dependent induction of tbx5a does not occur in the absence of the hoxba and hoxbb clusters.

We have moved the broader interpretation to the Discussion, where we now emphasize that our data are compatible with multiple models. One possibility is a linear pathway in which RA induces Hox expression that subsequently activates tbx5a. Alternatively, it is also plausible that RA induces Hox expression and that RA and Hox proteins act cooperatively to induce tbx5a. Our findings do not distinguish between these possibilities, and both models remain consistent with the data. We believe this restructuring addresses the reviewer’s concern by keeping the Results factual and limiting mechanistic interpretation to the Discussion.

The authors have carried out a functional test for the function of hoxb6 and hoxb8 in the hemizygous hoxb mutant background. What is lacking is any expression analysis to demonstrate whether Hoxb6b or Hoxb8b are even expressed in the appropriate pectoral fin territory to be able to contribute to pectoral fin development, either in this assay or in normal pectoral fin development.

We thank the reviewer for emphasizing the importance of expression analyses. In response, we performed a comprehensive whole-mount in situ hybridization survey of all eight PG4–8 Hox genes from the hoxba and hoxbb clusters (hoxb4a, hoxb5a, hoxb5b, hoxb6a, hoxb6b, hoxb7a, hoxb8a, and hoxb8b) during pectoral fin development (18–30 hpf). Among these, only hoxb4a, hoxb5a, and hoxb5b displayed detectable expression in the developing pectoral fin buds. In contrast, hoxb6a, hoxb6b, hoxb7a, hoxb8a, and hoxb8b were not expressed in this territory. These new data have been incorporated into the revised manuscript (Fig. 7M-R). We believe that this dataset provides a more complete and systematic picture of which Hoxb genes are available to function in pectoral fin development, and we are grateful to the reviewer for this valuable suggestion, which significantly strengthened our study.

(The term "compensate" used in this section is confusing/misleading.)

We thank the reviewer for this helpful remark. We agree that the term “compensate” was misleading in this context, as it could be confused with genetic compensation mechanisms such as transcriptional adaptation. To avoid this ambiguity, we have revised the wording.

Specifically, we replaced “compensate for” with “mimic the effect of” or “phenocopy” depending on the context. We believe this revision improves clarity and prevents misunderstanding.

The authors' confounding results described in Figures 6-7 are consistent with the challenges faced in other model organisms in trying to explore the function of genes in the hox cluster and the known redundancy that exists across paralogous groups and across individual clusters. Given the experimental challenges in deciphering the actual functions of individual or groups of hox genes, a discussion of the normal expression pattern of individual and groups of hox genes (and how this may change in different mutant backgrounds) could be helpful to make conclusions about likely normal function of these genes and compensation/redundancy in different mutant scenarios.

We appreciate the reviewer’s thoughtful comment. We agree that functional analyses of Hox genes are often complicated by redundancy within and across clusters. In this revision, we have included additional expression data of PG4–8 genes from the hoxba and hoxbb clusters, showing that only hoxb4a, hoxb5a, and hoxb5b are expressed in the fin buds. Although we did not analyze expression changes across mutant backgrounds in this study, we consider this an important direction for future experiments.

Reviewer #2 (Public review):

Summary:

The authors of this manuscript performed a fascinating set of zebrafish mutant analyses on hox cluster deletion and pinpointed the cause of the pectoral fin loss in one combinatorial hox cluster mutant of Hoxba and Hoxbb.

Strengths:

The study is based on a variety of existing experimental tools that enabled the authors' past construction of hox cluster mutants, and is well-designed. The manuscript is well written to report the authors' findings on the mechanism that positions the pectoral fin.

Weaknesses:

The study does not focus on the other hox clusters other than ba and bb, and is confined to the use of zebrafish, as well as the comparison with existing reports from mouse experiments.

We thank the reviewer for the thoughtful and encouraging evaluation of our manuscript. We are pleased that the strengths of our study design and clarity of writing were recognized. We also acknowledge the noted limitations, and while our focus here is on zebrafish hoxba and hoxbb clusters, we agree that future studies should expand to other hox clusters and additional models. Below, we provide individual responses to the specific points raised.

Reviewer #1 (Recommendations for the authors):

(1) Some additional expression analyses of Hoxb6/b8 etc, could be carried out to address some issues raised in the main review.

We thank the reviewer for this suggestion. In response, we performed additional whole-mount in situ hybridization analyses of PG4–8 genes from the hoxba and hoxbb clusters, including hoxb6b and hoxb8b. These experiments showed that only hoxb4a, hoxb5a, and hoxb5b are expressed in the developing fin buds, whereas hoxb6a, hoxb6b, hoxb7a, hoxb8a, and hoxb8b are not. We have incorporated these new data into the revised manuscript (Figure 7M-R), which we believe clarify why functional tests of hoxb6b and hoxb8b did not uncover specific requirements in fin development.

(2) The discussion section, particularly the more speculative section on evolutionary significance, could be reduced. Discussion of pelvic fin could be removed also, as this has not and could not be addressed with the current experimental design.

We thank the reviewer for this helpful suggestion. In line with the recommendation, we have reduced the speculative section on evolutionary significance in the Discussion to make it more concise and focused. We have also removed the discussion of pelvic fins, as these were not directly addressed by our current experimental design. We believe these changes improve the clarity and focus of the Discussion section.

(3) The conclusions on transformation to cardiac identity could be reevaluated and presented differently.

We appreciate the reviewer’s insightful comment. In the revised manuscript, we have toned down our interpretation regarding a transformation to cardiac identity. Instead, we now describe the findings more cautiously, emphasizing the clear loss of fin precursors rather than a definitive acquisition of cardiac fate. We believe this revision presents a more balanced interpretation of the data.

(4) Minor typographical - I would suggest removing "Genetic Evidence:" from the title.

We appreciate the reviewer’s suggestion. In accordance with this comment, we have revised the title to: “HoxB-derived hoxba and hoxbb clusters are essential for the anterior-posterior positioning of zebrafish pectoral fins”.

Reviewer #2 (Recommendations for the authors):

(1) The authors mention the redundancy (between the a type and b type) of Hox clusters derived from an additional whole genome duplication in the teleost fish lineage. But, they do not refer to whether the zebrafish Tbx5 ortholog has an additional copy. This information helps the readers' interpretation of the data presented. First of all, tbx5a suddenly appears on line 143 without introducing its relationship with Tbx5, which needs to be explained in a revised manuscript.

We thank the reviewer for highlighting this important point. In zebrafish, there are indeed two Tbx5 orthologs, tbx5a and tbx5b. In the revised manuscript, we have modified the text around line 124 to introduce tbx5a in the context of its orthology to Tbx5, ensuring that its appearance in the Results is clear to the readers.

(2) I did not readily get whether the limb/fin 'positioning' that the authors focus on in this study is 'anteroposterior' positioning, but not anything else. If it is what is meant, the word 'anteroposterior' should just be inserted at the first appearance of the word 'positioning'.

We thank the reviewer for pointing this out. Our study specifically addresses the anteroposterior positioning of paired appendages, that is, how the initial site of pectoral fin formation is defined along the anterior–posterior axis of the body. To clarify this, we have revised the text to insert the word “anteroposterior” at the first appearance of the term “positioning” in both the Abstract and Introduction (lines 26 and 53). We believe this change resolves the ambiguity and makes the focus of our study explicit.

(3) Figure 5B also shows the remarkable reduction of hoxc1a expression, which the authors do not mention at all. I wonder how this is explained and how the authors justify no remark on this throughout the manuscript.

We thank the reviewer for this insightful comment. As correctly noted, we did observe a marked reduction of hoxc1a expression in Figure 5B. However, based on our genetic analyses, we consider that the causal genes underlying the phenotype are most likely located in hoxba and hoxbb clusters. Therefore, although the change in hoxc1a expression is indeed a notable phenomenon, we did not emphasize it in the manuscript in order to maintain focus on the primary clusters responsible for the observed phenotype (lines 240-241). We agree that this point should be acknowledged, and we have now added a brief note in the Results to clarify our findings.

(4) Figure 1 consists of multiple panels (A-M) but lacks panel D.

We apologize for the oversight. We have corrected it.

(5) Line 85 - precise role -> exact role.

We have corrected it (line 95).

(6) Line 87 - the vertebrate class Actinopterygii & the class Sarcopterygii.

Thank the reviewer for pointing out. We have corrected it (line 98-99).

(7) Line 90 - homologous -> orthologous.

We have corrected it (line 102).

(8) Figure 5 - For interpretability of the data, I suggest writing 'Paralogous groups' on the top of the panels A and B, and 'Cluster' vertically on the left.

We thank the reviewer for this helpful suggestion. As recommended, we have added

“Paralogous groups” at the top of panels A and B, and “Clusters” vertically on the left side of Figure 5 to facilitate interpretation of the data.

(9) Some subheading titles are too long. They can be shortened into 'hoxb5a and -b5b expression in pectoral fin buds are RA-dependent' instead of 'Expression patterns of hoxb5a and hoxb5b in pectoral fin buds are dependent on RA', for example.

We appreciate the reviewer’s suggestion regarding the length of the subheading titles. In response, we have shortened the relevant subheadings in both the Results and Discussion sections to make them more concise while retaining their scientific meaning. For example, the subheading originally written as “Expression patterns of hoxb5a and hoxb5b in pectoral fin buds are dependent on RA” has been revised to “hoxb5a/b5b expression in pectoral fin buds is

RA-dependent.” Similar adjustments have been made to other subheadings throughout these sections. We believe these changes improve readability and consistency without altering the intended content.

(10) Line 408 - why tetrapods, instead of cartilaginous fishes, which are thought of as natural in this context?

We appreciate the reviewer’s careful reading and insightful comment. However, in response to Reviewer 1’s suggestion, we have substantially reduced the speculative section on evolutionary significance in the Discussion. As a result, this specific part of the text has now been deleted. We thank the reviewer for raising this point.

Version 2

Abstract

Vertebrate paired appendages, such as the pectoral fins in fish and the forelimbs in tetrapods, arise at specific regions along the anterior-posterior axis of the body. Hox genes have long been considered prime candidates for determining the anteroposterior positioning of these paired appendages during development. Evidence from various model organisms, including mouse and chick, supports a role for Hox genes in limb positioning. However, despite extensive phenotypic analyses of numerous single and compound Hox knockout mice, clear genetic evidence for substantial defects in limb positioning has been limited, leaving questions unresolved. In a previous study, we generated seven distinct hox cluster-deficient mutants in zebrafish. Here, we provide genetic evidence that zebrafish hoxba;hoxbb cluster-deleted mutants specifically exhibit a complete lack of pectoral fins, accompanied by the absence of tbx5a expression in pectoral fin buds. In these mutants, tbx5a expression in the pectoral fin field of the lateral plate mesoderm fails to be induced at an early stage, suggesting a loss of pectoral fin precursor cells. Furthermore, the competence to respond to retinoic acid is lost in hoxba;hoxbb cluster mutants, indicating that tbx5a expression cannot be induced in the pectoral fin buds. We further identify hoxb4a, hoxb5a, and hoxb5b as pivotal genes underlying this process. Although the frameshift mutations in these hox genes do not recapitulate the absence of pectoral fins, we demonstrate that deletion mutants at these genomic loci show the absence of pectoral fins with low penetrance. Our results suggest that, by establishing the expression domains along the anteroposterior axis, hoxb4a, hoxb5a, and hoxb5b within hoxba and hoxbb clusters cooperatively determine the positioning of zebrafish pectoral fins through the induction of tbx5a expression in the restricted pectoral fin field. Our findings also provide insights into the evolutionary origin of paired appendages in vertebrates.

Introduction

In jawed vertebrates, paired appendages—such as the pectoral and pelvic fins in fish, and their homologous forelimbs and hindlimbs in tetrapods—develop at precise locations along the anterior-posterior axis of each species. These paired appendages arise from progenitor cells located in distinct regions of the lateral plate mesoderm (Murata et al., 2010; Nishimoto and Logan, 2016; Shimada et al., 2013). The anteroposterior positioning of these paired appendages has long fascinated researchers, yet our understanding of the molecular mechanisms underlying this process remains limited.

In bilaterian animals, Hox genes—encoding evolutionarily conserved homeodomain-containing transcription factors—provide positional information and developmental timing along the anterior-posterior axis (Iimura and Pourquie, 2007; Izpisua-Belmonte and Duboule, 1992; Krumlauf, 1994). A defining feature of Hox genes is their structural organization into Hox clusters, where multiple Hox genes are arranged in a precise order. Additionally, Hox clusters exhibit a distinctive phenomenon known as Hox collinearity, where the genomic arrangement of Hox genes correlates with specific developmental regions along the body axes (Dolle et al., 1989; Duboule and Dolle, 1989; Graham et al., 1989). In vertebrates, Hox clusters underwent divergence due to two rounds of whole-genome duplication early in vertebrate evolution (Dehal and Boore, 2005; Ohno, 1970), leading to the establishment of four distinct Hox clusters (HoxA, HoxB, HoxC, and HoxD), each consisting essentially of 1-13 paralogous groups in tetrapods. In contrast, teleost fishes experienced an additional teleost-specific whole-genome duplication, followed by the loss of a hox cluster, resulting in seven hox clusters in zebrafish (Amores et al., 1998; Woltering and Durston, 2006).

Multiple studies in chick and mouse have suggested that the initial anteroposterior position of limbs is regulated by Hox genes (Tanaka, 2013; Tickle, 2015). For example, the anterior boundaries of Hox gene expression domains were shown to align with future limb positions in chick embryos (Burke et al., 1995; Cohn et al., 1997). Experimental manipulations in avian embryos, including overexpression or interference of specific Hox genes, led to altered positions of forelimb buds (Moreau et al., 2019). In mouse, Hoxb5 knockout mutants exhibit a rostral shift of forelimb buds with incomplete penetrance (Rancourt et al., 1995), while other subtle shifts in limb positioning have also been reported in Hox mutants (Royle et al., 2021). Furthermore, loss of Gdf11 in mice, which alters the expression of posterior Hox9–13 genes, causes posterior displacement of hindlimb buds, whereas ectopic Gdf11 induces anterior shifts of hindlimb position (Matsubara et al., 2017). At the molecular level, Hox proteins directly bind to the Tbx5 limb enhancer and regulate its expression, providing a mechanistic link between Hox activity and forelimb initiation (Minguillon et al., 2012; Nishimoto et al., 2014). Together, these studies provide strong evidence across different model organisms that Hox genes contribute to the regulation of limb positioning. Nevertheless, despite the generation of numerous single and compound Hox knockout mice, no severe defects in the initial positioning of limb buds have been documented, and the precise mechanisms by which Hox genes specify limb position remain incompletely understood. This stands in contrast to the well-established role of Hox genes in limb patterning after limb bud formation, where paralogous group 9–13 genes in the HoxA and HoxD clusters cooperatively control the proximal–distal axis of developing limbs (Boulet and Capecchi, 2004; Davis et al., 1995; Fromental-Ramain et al., 1996a; Fromental-Ramain et al., 1996b; Kmita et al., 2005). Consequently, the exact role of Hox genes in defining the initial position of limb formation along the anterior-posterior axis remains unclear.

Using zebrafish, which belong to the vertebrate class Actinopterygii and are distinct from the tetrapods of the class Sarcopterygii, we previously generated mutants lacking each of the seven zebrafish hox clusters using the CRISPR-Cas9 method (Yamada et al., 2021). In our recent genetic analyses of hoxaa, hoxab, and hoxda clusters—orthologous to the mouse HoxA and HoxD clusters—we demonstrated that, similar to mice, these zebrafish hox clusters cooperatively play an essential role in the formation of the pectoral fins (Ishizaka et al., 2024), which are homologous to the forelimbs. In this study, we provide the first genetic evidence, to our knowledge, that Hox genes specify the positions of paired appendages in vertebrates. The double-deletion mutants of hoxba and hoxbb clusters, derived from the ancient HoxB cluster, exhibit a complete absence of pectoral fins due to the failure to express tbx5a. Despite incomplete penetrance, we propose a model in which hoxb4a, hoxb5a, and hoxb5b cooperatively provide positional cues along the anterior-posterior axis within the lateral plate mesoderm, thereby specifying the initial positions for fin bud formation through the induction of tbx5a in the pectoral fin field.

Results

Absence of pectoral fins in hoxba;hoxbb cluster-deleted zebrafish

In a previous study, we created seven individual hox cluster-deficient mutants in zebrafish using the CRISPR-Cas9 system (Yamada et al., 2021). Among these mutants, hoxba cluster-deleted embryos exhibited morphological abnormalities in their pectoral fins at 3 dpf (Figure 1A-C). To further investigate the pectoral fin phenotype in hoxba cluster mutants, we first analyzed the expression patterns of Tbx5 orthologs. Zebrafish possess two paralogs. tbx5a and tbx5b (Boyle-Anderson et al., 2022). Among them, tbx5a plays a predominant role in the initial induction of pectoral fin buds in zebrafish (Ahn et al., 2002; Garrity et al., 2002; Ng et al., 2002). When comparing with tbx5a expression in wild-type embryos, we found that the tbx5a signal was reduced in the pectoral fin buds of hoxba cluster mutants (Figure 1G, H). Given that hoxba and hoxbb clusters originated from the ancestral HoxB cluster through teleost-specific whole-genome duplication (Amores et al., 1998), there may be functional redundancy between them. Surprisingly, we discovered that the simultaneous deletion of both hoxba and hoxbb clusters resulted in the complete absence of pectoral fins (Figure 1A, F). In contrast, pectoral fins were present in hoxba-/-;hoxbb+/- and hoxba+/-;hoxbb-/- mutants (Figure 1D, E), indicating that an allele from either hoxba or hoxbb cluster is sufficient for the pectoral fin formation. Moreover, all embryos lacking pectoral fins were identified as hoxba;hoxbb double homozygous mutants, with the expected penetrance (n=15/252; 5.9%) being consistent with predictions based on Mendelian genetics (1/16=6.3 %). Furthermore, while hoxba;hoxbb cluster double homozygous mutants are embryonic lethal around 5 dpf, we could not detect any trace of pectoral fin development. Alongside this phenotype, the expression of tbx5a was significantly reduced to nearly undetectable levels in hoxba;hoxbb cluster mutants at 30 hpf (Figure 1G-L). The absence of pectoral fins in zebrafish hoxba;hoxbb cluster mutants sharply contrasts with the results of a previous study, which showed that mice lacking all HoxB genes except for Hoxb13 of the HoxB cluster did not exhibit apparent abnormalities in their forelimbs (Medina-Martinez et al., 2000). Our genetic results suggest that the zebrafish hoxba and hoxbb clusters cooperatively play a crucial role in pectoral fin formation.

Lack of pectoral fins in hoxba;hoxbb cluster-deleted mutants.

(A-F) Dorsal views of live zebrafish larvae at 3 dpf, obtained from intercrosses between hoxba;hoxbb cluster-deficient hemizygous fish. Arrowheads indicate the positions of the pectoral fins. (G-L) Expression patterns of tbx5a in the pectoral fin bud (arrowhead) at 30 hpf. Dorsal views are shown, and the genotype of each specimen was determined. For each genotype, reproducibility was confirmed with at least three different specimens. Genotyping revealed that all the embryos lacking pectoral fins (n=15) were hoxba;hoxbb double homozygotes. Scale bars are 100 μm.

Pectoral fin loss specific to hoxba;hoxbb cluster-deleted zebrafish

Multiple genetic studies in mice have demonstrated the functional redundancy of Hox genes across the four Hox clusters (Horan et al., 1995; McIntyre et al., 2007; van den Akker et al., 2001; Wellik and Capecchi, 2003; Wellik et al., 2002). To investigate whether the absence of pectoral fins is specific to hoxba;hoxbb cluster mutants, we created double-hox cluster mutants by crossing hoxba cluster mutants with six other hox cluster mutants in zebrafish. The combinatorial mutations of hoxba cluster with five other hox clusters, excluding hoxbb cluster, did not enhance the reduced expression of tbx5a (Figure 2A-H). Furthermore, we examined combined mutants of hox clusters, excluding hoxba and hoxbb. In hoxca-/-;hoxcb-/- cluster mutants, where the functions of HoxC genes are completely absent, the expression patterns of tbx5a in fin buds were indistinguishable from those of wild-type embryos (Figure 2I). Additionally, our previous study showed that tbx5a expression is unaffected in hoxaa-/-;hoxab-/-;hoxda-/- cluster mutants (Ishizaka et al., 2024), which lack all HoxA and HoxD-related genes. The results of our deletion mutations of zebrafish hox clusters in various combinations emphasize that the diminished expression of tbx5a in the pectoral fin buds is evident only in hoxba;hoxbb cluster mutants. Taken together, in contrast to the functional redundancy of the four Hox clusters in mice, our genetic analysis indicates that the Hox clusters responsible for the specification of pectoral fin buds are restricted to the HoxB-derived hoxba and hoxbb clusters among the seven hox clusters in zebrafish.

Significantly decreased expression of tbx5a in the pectoral fin buds is specific to hoxba;hoxbb cluster-deleted mutants.

(A-I) Expression patterns of tbx5a in the pectoral fin buds of combinatorial deletion mutants of zebrafish hox clusters. Dorsal views of embryos at 30 hpf are displayed. After capturing images, the genotype of each specimen was determined. For each genotype, reproducibility was confirmed with at least three different specimens. Scale bars are 100 μm.

Loss of fin progenitors with posterior expansion of cardiac marker expression

To understand the absence of pectoral fin formation in hoxba;hoxbb cluster mutants, we examined the expression patterns of tbx5a during embryogenesis (Figure 3A-F). In wild-type embryos at the 10-somite stage, tbx5a expression appeared as bilateral symmetric stripes in the anterior lateral mesoderm, extending from the posterior midbrain to the second somite (Figure 3A) (Ahn et al., 2002). These tbx5a-positive cells subsequently divided into two groups: the anteriorly migrating cells form the heart primordia, while the posteriorly migrating cells give rise to the future pectoral fin buds (Figure 3C, E). In contrast, abnormal expression patterns of tbx5a were evident at early stages in hoxba;hoxbb cluster mutants. Compared to wild-type embryos, the bilateral stripes of tbx5a expression were shortened along the anterior-posterior axis, and the tbx5a-positive signal was absent from the posterior region where pectoral fin progenitor cells typically emerge (Figure 3B). Although anterior migration toward the presumptive heart occurred, posterior migration toward the fin buds was undetectable, consistent with the loss of pectoral fin progenitor cells in hoxba;hoxbb cluster mutants (Figure 3D, F). These results underscore the early role of the hoxba and hoxbb clusters in the formation of pectoral fin buds through the induction of tbx5a expression. In zebrafish, studies have shown a reciprocal relationship between pectoral fin and heart progenitor populations in the anterior lateral mesoderm, where retinoic acid signaling promotes pectoral fin formation while restricting cardiac progenitors (Waxman et al., 2008). Therefore, we also analyzed cardiac development (Figure 3G-J). By examining the expression patterns of nkx2.5, an early cardiac progenitor marker (Chen and Fishman, 1996; Serbedzija et al., 1998), we found that the expression domain of nkx2.5 in hoxba;hoxbb cluster mutants extended more posteriorly compared to wild-type embryos (Figure 3G, H). Furthermore, the region positive for the differentiated cardiac cell-specific myosin light chain (cmlc2) showed abnormal morphology relative to wild-type embryos (Figure 3I, J) (Yelon et al., 1999). Taken together, these results indicate that hoxba and hoxbb clusters are essential for establishing the pectoral fin field in zebrafish. In their absence, pectoral fin progenitor cells fail to emerge, accompanied by a posterior expansion of cardiac marker expression domain. Our results highlight a shift in the balance between fin and cardiac progenitor populations within the anterior lateral plate mesoderm.

Loss of fin progenitors with posterior expansion of cardiac marker expression in hoxba;hoxbb mutants.

(A-F) Expression patterns of tbx5a were compared between sibling wild-type and hoxba;hoxbb homozygous embryos during embryogenesis. The range of tbx5a expression in the lateral mesoderm is indicated by a bracket. The progenitor cells of pectoral fins are indicated by an arrowhead. (G, H) Expression patterns of nkx2.5 were compared between sibling and hoxba;hoxbb mutants. Enlarged views are shown in (G’, H’). The different regions of nkx2.5 expression between wild-type and mutants are indicated by brackets. (I, J) Expression patterns of cmcl2 are shown. All images were captured from the dorsal side. For each stage, reproducibility was confirmed with at least three different specimens. Scale bars indicate 100 μm.

hoxba/bb clusters required for RA-mediated tbx5a induction in the fin buds

Retinoic acid (RA) is a well-known upstream regulator of Hox genes in vertebrates (Boncinelli et al., 1991; Langston and Gudas, 1994; Marshall et al., 1996). Loss of function of retinaldehyde dehydrogenase 2 (raldh2), an essential regulator in RA synthesis, leads to the absence of pectoral fins in zebrafish (Begemann et al., 2001; Grandel et al., 2002), a phenotype similar to that observed in hoxba;hoxbb cluster mutants. Previous studies have shown that exogenous RA treatments can rescue pectoral fin formation in raldh2-/- embryos (Gibert et al., 2006; Grandel et al., 2002). It has been suggested that RA directly or indirectly induces tbx5a expression in zebrafish fin buds (Gibert et al., 2006; Grandel and Brand, 2011; Neto et al., 2012). The phenotypic similarities between hoxba;hoxbb cluster mutants and raldh2 mutants prompted us to investigate whether RA exposure could also rescue the absence of pectoral fins in hoxba;hoxbb cluster mutants. To explore this, we introduced a frameshift mutation in raldh2 using CRISPR-Cas9, confirming the phenotypic recapitulation of previously described raldh2-/- mutants (figure-supplement 1). Consistent with prior results (Gibert et al., 2006; Grandel et al., 2002), RA treatment in our raldh2 mutants rescued the expression of tbx5a in pectoral fins (Figure 4A-D). In contrast, RA did not induce tbx5a expression in the fin buds of hoxba;hoxbb cluster mutants (Figure 4E-H), and no trace of pectoral fin formation was observed in the treated embryos even at 5 dpf. Furthermore, the endogenous expression patterns of raldh2 in hoxba;hoxbb cluster mutants seem indistinguishable from those in wild-type embryos (Figure 4I, J), suggesting that the response to RA is absent in the fin buds of hoxba;hoxbb cluster mutants, while RA synthesis appears intact. Although a model proposing that RA directly induces tbx5a expression has been suggested (Gibert et al., 2006; Grandel and Brand, 2011; Neto et al., 2012), our results indicate that RA-dependent induction of tbx5a does not occur in the absence of hoxba and hoxbb clusters. Given that hoxba;hoxbb cluster mutants retain the other five intact hox clusters, our findings further support the notion that the hoxba and hoxbb clusters are specifically responsible for the RA-mediated specification of pectoral fins.

Zebrafish hoxba;hoxbb mutants lack a response to RA in fin buds.

(A-D) Exogenous RA exposure in raldh2 mutants can rescue tbx5a expression in fin buds. Arrowheads indicate the positions of the pectoral fin buds. (E-H) RA treatments in hoxba;hoxbb cluster-deleted mutants do not rescue tbx5a expression in fin buds. (I, J) Expression patterns of raldh2 were compared between sibling and hoxba;hoxbb mutants. For each genotype, reproducibility was confirmed with at least three different specimens. All images were captured from the dorsal side. Scale bars represent 100 μm.

hoxb5a/b5b expression in fin buds is RA-dependent

The absence of pectoral fins, commonly observed in hoxba;hoxbb cluster mutants and raldh2 mutants, suggests that candidate hox genes may be downregulated in raldh2 mutants. To investigate this, we performed RNA-seq analysis to compare the expression profiles between sibling and raldh2 homozygous embryos. As expected, we found that most of the 3’-located hox genes are downregulated in raldh2-/- embryos (Figure 5A, B, figure-supplement 2). Although hoxc1a showed the most pronounced reduction, our analysis focused on hoxba and hoxbb clusters. Notably, the levels of hoxb5b transcripts were significantly reduced, along with those of hoxb5a among hoxba genes, in raldh2 mutants. Previous studies have shown that zebrafish hoxb5a and hoxb5b are expressed in the lateral plate mesoderm corresponding to presumptive pectoral fin buds (Waxman et al., 2008). Additionally, knockout mice for Hoxb5, which is homologous to zebrafish hoxb5a and hoxb5b, only exhibit altered positioning of forelimbs (Rancourt et al., 1995), among various other Hox knockouts that have been generated. Therefore, we examined whether RA regulates the expression of hoxb5a and hoxb5b in fin buds. In wild-type embryos, hoxb5a expression was observed in the neural tube, somites, and lateral plate mesoderm (Figure 5C). However, in raldh2 mutants, hoxb5a expression was significantly reduced, notably in the lateral plate mesoderm, which differentiates into the fin buds, where it was reduced to the point of being undetectable (Figure 5D). Since RA treatments in raldh2 mutants can rescue pectoral fin development, we found that hoxb5a expression in raldh2 mutants is also rescued by RA exposure (Figure 5C-E). Regarding the expression of hoxb5b, which diverged from the same ancestral gene as hoxb5a, RA-dependent expression was also confirmed, mirroring the observations made with hoxb5a (Figure 5F-H). These results suggest that both hoxb5a and hoxb5b are involved in the formation of the pectoral fins in zebrafish.

Expression patterns of hox5a and hoxb5b are regulated by RA.

(A) Schematic representation of the 49 hox genes organized into seven hox clusters in zebrafish. hox genes in hoxba and hoxbb clusters are highlighted with orange. (B) Expression profiles of the 49 hox genes in wild-type and raldh2-/- embryos at the 20-somite stage, analyzed by RNA-seq. The average FPKM of each hox gene was compared between sibling and raldh2 mutants. The absence of specific hox genes is indicated by a slash. (C-H) Expression patterns of hoxb5a and hoxb5b in sibling wild-type, raldh2-/-, and RA-treated raldh2-/- embryos at the 10-somite stage. Arrowheads indicate the presumptive positions of pectoral fin buds. For each genotype, reproducibility was confirmed with at least three different specimens. All images are captured from the dorsal side. Scale bars represent 100 μm.

Genetic screen for hox genes specifying pectoral fin buds

To understand the molecular mechanisms underlying pectoral fin specification, we sought to identify hox genes essential for fin formation through a genetic approach. We anticipated that frameshift mutations in relevant hox genes would phenocopy the absence of pectoral fins observed in hoxba;hoxbb mutants. Since hoxbb cluster contains fewer hox genes than hoxba cluster (Figure 5A), we focused our investigation on hoxbb cluster. Although hoxb5b is expressed in the presumptive fin buds (Figure 5F), other hoxbb genes may also be involved. Therefore, we simultaneously injected three gRNAs targeting hoxb5b, hoxb6b, and hoxb8b into hoxba hemizygous mutants, expecting various germline mutations in these hoxbb genes in the injected fish (Figure 6A). Subsequently, F1 fish with a hemizygous deletion of hoxba cluster and mosaic mutations were crossed with hoxba;hoxbb hemizygous mutants. Among the progeny, several embryos lacking pectoral fins were identified. Following genotyping and DNA sequencing of the target hoxbb genes, we found that all embryos without pectoral fins possessed frameshift mutations in hoxb5b, with hoxba-/-;hoxbb+/- being common among them (Figure 6B-E, table-supplement 1, figure-supplement 3). Similarly, we injected each crRNA for hoxb6b and hoxb8b into hoxba+/- mutants and confirmed that neither frameshift mutation in hoxb6b nor hoxb8b in hoxba-/-;hoxbb+/- led to the loss of pectoral fins (Figure 6F, G, figure-supplement 3). These results suggest that the frameshift mutation of hoxb5b alone can phenocopy the deletion of hoxbb cluster, recapitulating the absence of pectoral fins in hoxba;hoxbb cluster mutants.

Screening for hox genes responsible for the zebrafish pectoral fin formation.

(A) Schematic representation of the genetic screening for hox genes involved in the specification of pectoral fins in zebrafish. (B-G) Dorsal views of live mutant larvae at 3 dpf, obtained during the screening. After capturing images, genotyping and DNA sequencing in target hoxb genes were conducted. For the mutants illustrated in (C-G), hoxb genes are shown with frameshift mutations introduced. Detailed information is provided in Supplementary Table S1. (H-J) Dorsal views of hoxba-/-;hoxb5b-/- and hoxba-/-;hoxb5b-/-;hoxb6b-/- larvae at 3 dpf. (K-M) Expression patterns of tbx5a in hoxba-/- ;hoxb5b-/- and hoxba-/-;hoxb5b-/-;hoxb6b-/- mutants at 30 hpf. Arrowheads indicate the presumptive positions of pectoral fin buds. All images are captured from the dorsal side. Scale bars represent 100 μm.

However, we encountered an unexpected result: hoxba-/-;hoxb5b-/- embryos, which have a homozygous frameshift mutation in hoxb5b and a homozygous deletion of hoxba cluster, exhibited severely truncated but visibly present pectoral fins (Figure 6H, I). After analyzing hundreds of embryos from several crossings, we did not find any hoxba-/-;hoxb5b-/- embryos without pectoral fins. Consistent with this, the expression of tbx5a was detectable in the fin buds of hoxba-/-;hoxb5b-/- embryos (Figure 6K, L), in contrast to its absence in hoxba;hoxbb cluster mutants (Figure 1L). To examine the potential contributions of other hoxbb genes, we observed the phenotypes of hoxba-/-;hoxb5b-/-; hoxb6b-/- embryos; however, these mutants did not exacerbate the abnormalities and still formed pectoral fins (Figure 6J, M).

The phenotype of hoxba-/-;hoxb5b-/- embryos indicates that a frameshift mutation in hoxb5b does not fully mimic the complete deletion of the entire hoxbb cluster, suggesting that other mechanisms may be involved. A novel genetic compensation mechanism known as transcriptional adaptation has been identified: mRNA containing a premature termination codon (PTC) can induce increased expression of structurally related genes, thereby compensating for the function of mutated genes (El-Brolosy et al., 2019; Rossi et al., 2015). To avoid producing transcripts with PTC (Sztal and Stainier, 2020), we generated a full locus-deleted allele of hoxb5b using CRISPR-Cas9 with two crRNAs targeting both ends of the target locus (figure-supplement 4). In contrast to the phenotype of hoxba-/-;hoxb5b-/-, we found that hoxba-/-;hoxb5b de/dell mutants lack pectoral fins and exhibit a significant reduction in tbx5a expression when intercrossing hoxba+/-;hoxb5b+/del fish (Figure 7A, B, D, E). However, the occurrence rate of embryos lacking pectoral fins (n=3/120; 2.5 %) was lower than predicted based on Mendelian genetics (1/16; 6.3 %). Other hoxba-/-;hoxb5bdel/del embryos did exhibit shortened pectoral fins (Figure 7C). Additionally, concerning hoxba cluster, we also generated frameshift-induced hoxb5a-/-;hoxb5b-/-and a full locus-deleted hoxb5adel/del;hoxb5bdel/delembryos (figure-supplement 4). hoxb5a-/-;hoxb5b-/-larvae consistently displayed normal pectoral fins (Figure 7F). In contrast, hoxb5adel/del;hoxb5bdel/del mutants exhibited a slightly more pronounced phenotype of shortened pectoral fins, but no instances of missing pectoral fins were identified (Figure 7G). To further clarify which Hox genes could potentially contribute to fin bud formation, we analyzed the expression of PG4 and PG6–8 genes from hoxba and hoxbb clusters using whole-mount in situ hybridization. Among these, only hoxb4a showed detectable expression in the fin buds, whereas hoxb6a, hoxb6b, hoxb7a, hoxb8a, and hoxb8b were not detectable (Figure 7M-R). Furthermore, due to the functional redundancy observed between hoxb4a and hoxb5a observed in zebrafish vertebral patterning and the RA-dependent expression of hoxb4a (Grandel et al., 2002; Maeno et al., 2024), we generated frameshift-induced hoxb4a-/-;hoxb5a-/-;hoxb5b-/-embryos and an allele lacking the entire genomic region encompassing hoxb4a and hoxb5a (figure-supplement 4). Although hoxb4a-/-;hoxb5a-/-;hoxb5b-/- embryos did not lack pectoral fins (Figure 7H), some hoxb4a-b5adel/del;hoxb5b del/del embryos did lack pectoral fins, although at low penetrance (Figure 7I, n=3/397; 0.7%). The majority of hoxb4a-b5adel/del;hoxb5b del/del embryos displayed truncated pectoral fins (Figure 7J). The lack of pectoral fins was not detected in hoxb4a-b5adel/del and hoxb4adel/del;hoxb5bdel/del mutants (Figure 7K, L, figure-supplement 4). These results suggest that hoxb4a, hoxb5a, and hoxb5b may cooperatively contribute to the anterior-posterior positioning of pectoral fin buds in zebrafish. Furthermore, our findings reveal that transcriptional adaptation may partially play a role in genetic compensation for frameshift-induced hox mutations during pectoral fin formation, while also suggesting the potential involvement of other unknown mechanisms or additional genomic regions in the initial induction of zebrafish pectoral fin buds.

hoxb4a-b5adel/del;hoxb5bdel/del larvae partially recapitulate the absence of the pectoral fins.

(A-C) Dorsal views of zebrafish hoxba-/-;hoxb5bdel/del larvae at 3 dpf. (D, E) Expression patterns of tbx5a in sibling wild-type and hoxba-/-;hoxb5bdel/del (n=4) at 30 hpf. Arrowheads indicate the presumptive positions of pectoral fin buds. (F, G) Dorsal view of frameshift-induced hoxb5a-/-;hoxb5b-/- and hoxb5a del/del;hoxb5bdel/del larvae. (I-K) Dorsal view of frameshift-induced hoxb4a-/-;hoxb5a-/-;hoxb5b-/- and hoxb4a-b5a del/del;hoxb5bdel/del larvae. All images are captured from the dorsal side. Arrowheads indicate the presumptive positions of pectoral fin buds. (M-R) Expression patterns of zebrafish hoxb4a, hoxb6a, hoxb6b, hoxb7a, hoxb8a, and hoxb8b at the 10-somite stage. Dorsal views. Arrowheads indicate the presumptive positions of pectoral fin buds. Scale bars represent 100 μm.

Discussion

Since the establishment of gene targeting (Mansour et al., 1988), numerous Hox knockout mice have been generated. However, neither single nor compound Hox mutants, nor Hox overexpression approaches, have been reported to cause significant defects in limb positioning (Jurberg et al., 2013; Moreau et al., 2019). Consequently, the role of Hox genes in the positioning of vertebrate paired appendages remains unclear due to a lack of direct genetic evidence. In this study, we provide genetic evidence using zebrafish, demonstrating that specific combinatorial deletions of hoxba and hoxbb clusters result in the absence of pectoral fins and a lack of induction of tbx5a expression in the pectoral fin field. Based on our results, we propose a model in which Hox expressions in the lateral plate mesoderm, regulated by RA synthesized in the paraxial mesoderm, provide positional information and induce tbx5a expression in the presumptive fin buds. Our genetic results shed light on longstanding fundamental questions regarding the molecular mechanisms by which the positioning of paired appendages is established in vertebrates.

hoxba/bb clusters (possibly hoxb4a, b5a, b5b) define the pectoral fin field

We showed that hoxba;hoxba mutants exhibit a lack of pectoral fins. Abnormal expression of tbx5a in the lateral plate mesoderm, where pectoral fin precursor cells normally arise, was evident from its initial expression in the mutants. We presume that hox genes in hoxba and hoxbb clusters, induced by RA, define the region along the anterior-posterior axis where pectoral fin formation can occur and promote the expression of tbx5a in the lateral plate mesoderm. Importantly, tbx5a expression in the fin bud was not induced even by RA exposure in hoxba;hoxbb mutants (Figure 4E-H). In zebrafish, it has been suggested that RA synthesized in the paraxial mesoderm acts directly on the lateral plate mesoderm, or that RA induces tbx5a expression via wnt2b in the intermediate mesoderm (Gibert et al., 2006; Grandel and Brand, 2011; Neto et al., 2012). However, our results suggest that intact hoxba and hoxbb clusters are required for RA-dependent induction of tbx5a expression. One possible explanation is a linear pathway in which RA induces hox expression, which then activates tbx5a. Alternatively, RA may induce hox expression and, together with Hox proteins, act cooperatively to activate tbx5a. Our data are consistent with both models. Through genetic studies, albeit with low penetrance, we further showed that hoxb4a, hoxb5a, and hoxb5b play a significant role in the induction of pectoral fin buds. The homologous Hoxb4 and Hoxb5 proteins in mice have been shown to bind directly to the enhancer of Tbx5 (Minguillon et al., 2012). Moreover, in chickens, Tbx5 expression in the forelimb field is restricted by posterior Hox genes, preventing posterior expansion (Moreau et al., 2019; Nishimoto et al., 2014). In contrast, our combinatorial deletions of zebrafish hox clusters did not produce posterior expansion of tbx5a expression. We therefore propose that, in zebrafish, hoxb4a, hoxb5a, and hoxb5b, induced by RA in the paraxial mesoderm, act primarily to promote tbx5a expression in the pectoral fin field.

Partial phenocopy of locus-deleted hox mutants implies other unidentified mechanisms

In this study, we observed that hox mutants with deletions of genomic loci exhibited more pronounced pectoral fin phenotypes than frameshift-induced hox mutants. The frameshift mutants used in this study are likely to be loss-of-function alleles, as we demonstrated several defects in vertebral patterning in these mutants (Maeno et al., 2024). These results suggest that transcriptional adaptation may occur during the formation of pectoral fins in frameshift-induced hox mutants, potentially compensating for the loss of function. In our previous studies, which focused on dorsal and anal fin patterning, vertebral patterning, and swim bladder formation, we demonstrated that the phenotypes observed in zebrafish hox cluster-deleted mutants could essentially be replicated by introducing frameshift mutations into hox genes within the cluster (Adachi et al., 2024; Maeno et al., 2024; Satoh et al., 2024; Yamada et al., 2021). Therefore, creating a mutant with a deleted locus that does not induce transcriptional adaptation was unnecessary. One factor that may explain the differences in results between this study and our prior findings is that even a small amount of hox gene products essential for pectoral fin positioning may be sufficient to induce fin buds, as shown in Figure 1, potentially leading to the manifestation of gene compensation through transcriptional adaptation in frameshift-induced mutants. Alternatively, transcriptional adaptation may function differently depending on various developmental processes or stages. Interestingly, all the aforementioned phenotypes reproducible by frameshift mutations were observed at later developmental stages than pectoral fin formation. Additionally, in hox locus-deleted mutants, only some hoxba-/-;hoxb5b de/dell or hoxb4a-b5adel/del;hoxb5b del/del embryos were able to reproduce the loss of the pectoral fin, suggesting the presence of other compensatory mechanisms. The organized expression patterns of Hox genes are thought to be regulated by complex regulatory controls, with their expression regulated according to the genomic order within the cluster. In the absence of Hox function, neighboring intact Hox genes could compensate for the loss of function. It is also possible that unidentified non-coding regions in hoxba and hoxbb clusters are involved in pectoral fin positioning. These issues should be clarified in future studies.

Hox genes and the evolutionary origin of pectoral fins

According to fossil records from the Cambrian period, the earliest vertebrates did not possess paired fins (Janvier, 1996; Shubin et al., 1997). Later in evolution, vertebrates acquired primitive pectoral fins, which served as evolutionary precursors to tetrapod forelimbs, before diverging into ray-finned and lobe-finned fishes. Our results align well with this evolutionary scenario. Hox genes, which are conserved across bilaterian animals, were already present in primitive vertebrates without paired fins, suggesting that their functional diversification may have facilitated the origin of new appendages. The early fossils of vertebrates with pectoral fins were discovered in Ordovician and Silurian jawless fishes (Janvier, 1996; Shubin et al., 1997). Our finding that hoxba;hoxbb mutants completely lack pectoral fins implies that molecular evolution within the ancestral HoxB cluster contributed to the emergence of pectoral fins. Although the precise molecular modifications remain unclear, one plausible model is that HoxB genes acquired the capacity to induce Tbx5 expression in the lateral plate mesoderm, establishing the RA–HoxB–Tbx5 gene network essential for fin initiation. We presume that the acquisition of pectoral fins was driven by multiple genomic changes accumulated over time, among which molecular evolution within the HoxB cluster may have represented one of the principal factors contributing to the origin of this novel appendage.

The positioning of pectoral fins is restricted to specific hox clusters in zebrafish, in contrast to the functional overlap seen in mice

Taking advantage of our availability of seven individual hox cluster mutants in zebrafish, we generated multiple cluster deletions and demonstrated that hox genes required for establishing pectoral fin buds are restricted to hoxba and hoxbb clusters. In contrast, analyses of knockout mice have revealed extensive functional overlap among the four Hox clusters. For instance, mice lacking all HoxB genes except for Hoxb13 do not lose forelimbs (Medina-Martinez et al., 2000), and forelimbs are still present in Hoxa5;Hoxb5;Hoxc5 triple knockout mice, which lack all Hox genes of the paralog group 5 (Xu et al., 2013). Thus, while zebrafish hoxba;hoxbb mutants exhibit a striking phenotype of absent pectoral fins, forelimbs are preserved in comparable mouse mutants. We speculate that paired fins emerged after the quadruplication of primitive Hox clusters, and that novel functions related to pectoral fin formation arose within the HoxB cluster. In zebrafish, these functions remain confined to hoxba and hoxbb clusters, making them indispensable for pectoral fin development. Interestingly, a similar cluster-specific confinement has been observed in median fins: in both zebrafish and medaka, Hox genes responsible for positioning dorsal and anal fins are primarily located in HoxC-related clusters (Adachi et al., 2024). In contrast, in mice, functional redundancy among Hox clusters—likely accumulated during over half a billion years of vertebrate evolution—may mask the requirement of HoxB genes for forelimb formation. While the precise evolutionary mechanisms remain to be clarified, our results suggest that cluster-specific specialization in teleosts contrasts with broader redundancy across Hox clusters in mammals.

Materials and methods

Zebrafish

Riken Wild-type (RW) zebrafish, obtained from the National BioResource Project in Japan, were maintained at 27°C with a 14-hour light/10-hour dark cycle. Embryos were collected from natural spawning, and the larvae were raised at 28.5°C. Developmental stages of the embryos and larvae were determined based on hours post-fertilization (hpf), days post-fertilization (dpf), and developmental stage-specific features (Kimmel et al., 1995; Parichy et al., 2009). The alleles of hox cluster-deleted mutants used in this study were previously generated using CRISPR-Cas9 (Yamada et al., 2021). The following frameshift-induced hox mutants were created using CRISPR-Cas9: hoxb5asud125, hoxb5b sud136, hoxb6b sud137, and hoxb4a;hoxb5a sud144 (Maeno et al., 2024). All experiments involving live zebrafish were approved by the Committee for Animal Care and Use of Saitama University.

Generation of mutants by the CRISPR-Cas9 system

All the mutants used in this study were generated using the Alt-R CRISPR-Cas9 system (Integrated DNA Technologies). To create frameshift-induced mutants, gene-specific crRNAs were incubated with common tracrRNA, followed by Cas9 nuclease. For the generation of mutants that delete large genomic regions, two crRNAs targeting both ends of the regions were incubated with common tracrRNA and Cas9 nuclease. Approximately one nanoliter of the crRNA:tracrRNA-Cas9 RNA-protein complex was injected into fertilized embryos, which were subsequently raised to juvenile fish. Candidate founder fish were selected using a heteroduplex mobility shift assay for frameshift mutants (Ota et al., 2013) and by amplifying genomic deletions with specific primers. After sexual maturation, candidate founder fish were mated with wild-type fish to produce heterozygous F1 offspring. Among the F1 offspring, mutant fish carrying the same mutation were identified through PCR-based genotyping followed by DNA sequencing. The target-specific sequences of the crRNAs used in this study are listed (table-supplement 2). The frozen sperm from the mutants used in this study have been deposited in the National BioResource Project Zebrafish in Japan (https://shigen.nig.ac.jp/zebra/) and are available upon request.

Genotyping of mutants

For the phenotypic analysis of embryos and the maintenance of mutant fish lines, PCR-based genotyping was performed. Briefly, genomic DNA was extracted from the analyzed embryos or the partially dissected caudal fins of anesthetized larvae or fish using the NaOH method (Meeker et al., 2007), which served as the template for PCR. The sequences of the primers used for genotyping frameshift-induced mutants are listed in the supplementary Table 3. For genotyping deletion mutants of hox genes, PCR was conducted using a combination of three primers, with their sequences also provided (table-supplement 3). Genotyping of hox cluster-deleted mutants was performed using PCR as previously described (Yamada et al., 2021). After the reactions, the PCR products were separated by electrophoresis. Based on differences in the lengths of PCR products derived from wild-type and mutated alleles, either a 2 % agarose gel in 0.5x TBE buffer, a 15 % polyacrylamide gel in 0.5x TBE buffer, or direct DNA sequencing was utilized to determine the genotype.

Whole-mount in situ hybridization

Whole-mount in situ hybridization was carried out as described (Thisse and Thisse, 2014). After the staining, the embryos mounted in 70-80 % glycerol were captured under the stereomicroscope (Leica M205 FA) with a digital camera (Leica DFC350F). After taking images, PCR-based genotyping was carried out as described above.

RNA-seq analysis

For RNA-seq analysis, embryos were obtained by the intercrosses between raldh2 heterozygous fish. At the 18-20-somite stages, sibling and homozygous embryos were separated based on the phenotype, and 20 embryos per tube were collected. Then, total RNA was extracted by ISOGENE (Nippon Gene) and followed by DNase I treatment (Takara). After the quality check of the isolated RNA, RNA-seq analyses (n=2 for each) were carried out by Genewiz, Azenta Life Science.

Treatment of retinoic acid in the embryos

Treatment of retinoic acid on the embryos was performed as previously described, with minor modifications (Gibert et al., 2006; Grandel et al., 2002). Briefly, embryos were obtained from intercrosses between raldh2 heterozygous mutants. At the 70 % epiboly stage, the embryos were soaked in a solution containing DMSO or 10-8M RA and incubated at 28.5°C. Embryos were fixed at 30 hpf for analysis of tbx5a expression and at the 10-somite stage for examination of hox gene expression.